---
title: "Age Structure Population Modelling"
output:
  slidy_presentation:
    highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
  beamer_presentation:
    highlight: pygments
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sensitivity)
library(tidyverse)
library(purrr)
library(ggpubr)
```

# See notes on population matrix modeling (agestructure.pdf)
# Set up the Leslie Matrix

[Leslie Matrix](lecture17_matrix/SimpleLeslie.pdf)

```{r setupleslie}
nclasses = 5

# create a growth matrix to store fertility and survivability information
gmatrix=matrix(nrow=nclasses, ncol=nclasses)
gmatrix

# change NAs to zero
gmatrix[]=0.0
gmatrix

# assign values for fertility for each of the age classes 
# notes that fertility rates are births per capita (number in class) per
# time step - time step here is decade - but could be hours
fert =  c(0,0.4,0.3,0.1,0.01)

# enter into our matrix
 gmatrix[1,]=fert
 
 # now add survivability 
 # survivability (to the next class) is also per time step
gmatrix[2,1]=0.5
gmatrix[3,2]=0.8
gmatrix[4,3]=0.2
gmatrix[5,4]=0.1

# we als want to to account for the oldest population group - they don't transfer to another group
# but they do die - this will be survivability per time step but they just stay in their class/group
gmatrix[5,5]=0.1

gmatrix

```

# Use Leslie matrix to  grow the population


```{r simplegrowth}

# start with an initial population, lets say 1 individual in each class

p0 = rep(1, times=nclasses)

# advance to the next time step
# note the use of matrix multiplication

p1 = gmatrix %*% p0
p1

# has the total number of individuals changed?




sum(p1)
sum(p0)

# growth rate
sum(p1)/sum(p0)

#add another year
p2 = gmatrix %*% p1

# combined
pop = cbind.data.frame(p0,p1,p2)
pop$age = c("0-1","1-10","11-20","21-50","50 and above")

popl = pop %>% pivot_longer(names_to="timestep",values_to="pop",-age)
ggplot(popl, aes(timestep, pop,fill=as.factor(age)))+geom_col(position="dodge")+labs(fill="Age Group")


# how might you change parameters to have the population grow?

```

# Multiple Time Steps

use function to evolve a population through time

* inputs = survivability, fertility, initial population, time steps
* output = final population matrix
* a dynamic model - difference equations - similar to our diffusion model
 
```{r multitime}

source("../R/evolve_pop.R")

# example evolve China pop
# fertility rates
F1 = 0.0
F2 = 0.25/2
F3 = 0.5/2
F4 = 0.25/2

# survivability - based on mortality rates per thousand per decade
p12 = 1-79/1000 
p23 = 1-9.8/1000
p34 = 1-16/1000
p45 = 1-25/1000
p56 = 1-10/1000
p67 = 1-110/1000
p78 = 1-270/1000
p89=1-630/1000
p99 = 0.1

# initial population parameters
ini = c(235,224,182,124,95,69,42,24,6)
ndecades = 15
fert_china = c(0,F2,F3,F4,0,0,0,0,0)
surv_china = c(p12,p23,p34,p45,p56,p67,p78,p89,p99)
china_pop=evolve_pop(fert_china, surv_china, ini, ndecades)

head(china_pop)

# graph diffeernt components of the output
# total population

# add decade 
decade = seq(from=1, to=ndecades)
china_tot = cbind.data.frame(decade=decade, poptot=china_pop$poptot)
ggplot(china_tot, aes(decade, poptot))+geom_col()+labs(y="Total Popultation")


# plot information about ages
china_ages = cbind.data.frame(decade=decade, t(china_pop$popbyage))
china_agesl = china_ages %>% pivot_longer(names_to="agecat", values_to="pop",-decade)
ggplot(china_agesl, aes(decade, pop, fill=agecat))+geom_col()+labs(y="Population", fill="Age Group")

```
 
#  Find parameters that mantain a stable population!
 
# Microbial Populations - Another example 
 
Lets try with a more complicated example - evolving an age structured population

Sensitivity **To**: fertility and survivability rates
  
Sensitivity **Of**: total population after 3 months
  
Assume we are looking at microbial population and monthly fertility and mortality rates; but format the same as china example

# Microbial Population Parameters

These microbes live for approximately 2 years
they are reproduce after 1 month of maturation

* 3 age groups
* 0-1 months old
* 1-2 months old
* more than 2 months

we can impact their survivability so we want to know how sensitive populations will be after 6 months if we do something to alter survivablity rates; 
but we also have uncertainty in what we know about reproduction rates


* 0-1 months - no fertiliaty = survivability about 90%
* 1-2 months - population tends to double, survivability about 70%
* more 2 months - population fertility around 20%, suvivability only 30%


Try setting this up...
```{r  example}

source("../R/evolve_pop.R")


# set up best guess case
# survivability - based on mortality rates per thousand per decade
p12 = 0.9
p23 = 0.7
p34 = 0.2

F1= 0.0
F2 = 2.0
F3 = 0.3

# initial population parameters
ini = c(1000,1000, 1000)
nmonths = 12
fert_micro = c(F1,F2,F3)
surv_micro = c(p12,p23,p34)
micro_pop=evolve_pop(fert_micro, surv_micro, ini, nmonths)

# population after 12 months
micro_pop$poptot[12]
```

# Sensitivity Analysis for matrix populations

Now lets use Sobel sensitivity analysis to explore the how our intervention might impact the population...


AND account for uncertainty in some of our parameters

Reminder
Sobel sensitivity analysis breaks variance of output into 

* variance associated directly with parameter alone (fraction associated with each parameter, sum to 1)
* variance associated with parameter and interaction with other parameters (sum can be more than 1)

# Steps

* run sobel to get parameter sets in a sensitivity analysis object
* run model with those parameter sets
* tell the senstivity object about results associated with each parameter set
* look at sensitivity analysis metric from sobel

```{r sobel}

library(sensitivity)

# survivability - based on mortality rates per thousand per decade
nsample=200

p12 = 0.9
p23 = 0.7
p34 = 0.2

# create our two samples for Sobel
# first do our survivability
ps1 = cbind.data.frame(p12=runif(min=0.4, max=0.92, n=nsample), p23 = runif(min=0.3, max=0.72, n=nsample),
                       p34 = runif(min=0.05, max=0.22, n=nsample))

ps2 = cbind.data.frame(p12=runif(min=0.4, max=0.92, n=nsample), p23 = runif(min=0.3, max=0.72, n=nsample),
                       p34 = runif(min=0.05, max=0.22, n=nsample))

# now include uncertainty in our Fertility
F1= 0.0
F2 = 2.0
F3 = 0.3

fs1 = cbind.data.frame(f1=rnorm(mean=0, sd=0, n=nsample), 
                       f2 = rnorm(mean=2.0, sd=0.1, n=nsample),
                       f3 = rnorm(mean=0.3, sd=0.01, n=nsample))

fs2 = cbind.data.frame(f1=rnorm(mean=0, sd=0, n=nsample), 
                       f2 = rnorm(mean=2.0, sd=0.1, n=nsample),
                       f3 = rnorm(mean=0.3, sd=0.01, n=nsample))


# put servivability and fertility together
allp1 = cbind.data.frame(ps1,fs1)
allp2 = cbind.data.frame(ps2,fs2)


# get sobel samples

sens_micro = sobolSalt(model = NULL, allp1, allp2, nboot = 100)

head(sens_micro$X)
nsim=nrow(sens_micro$X)

# add parameter names
colnames(sens_micro$X)=colnames(allp1)
# run model and save what we care about: final population after 12 months 
# this is already output by evolve_pop so we don't need a compute_metric function

ini = c(1000,1000, 1000)
nmonths = 12


# as before combine our application of the the dynamics model - for each
# parameter set, with code to extract our metric of interest (final population)
p_wrapper = function(p12, p23, p34, f1,f2,f3,use_func, initialpop, nstep ) {
fertility=c(f1,f2,f3)
survivability= c(p12,p23,p34)
res = use_func(survivability =survivability, fertility = fertility, initialpop=initialpop, nstep=nstep)
# now return the final population total
return(finalpop=res$poptot[nstep])
}

# use pmap here so we can specify rows of our sensitivity analysis parameter object 
res = as.data.frame(sens_micro$X) %>% pmap_dbl(p_wrapper, initialpop=ini, nstep=nmonths, use_func=evolve_pop)
         
# plot results (variation in final population across all parameter)
# ggplot needs a dataframe - so do a quick conversion with data.frame
ggplot(data.frame(finalpop=res), aes(x=finalpop))+geom_density()

# or a boxplot
ggplot(data.frame(finalpop=res), aes(x="", y=finalpop/1000) )+geom_boxplot(fill="blue")+
  theme(axis.title.x = element_blank())+labs(y="Final Pop (in 1000s)")

# give our results to sensitivity structure

sens_micro=tell(sens_micro, res)

# loot at results
sens_micro$S
sens_micro$T

# make this a bit easier to see
rownames(sens_micro$S) = colnames(sens_micro$X)
sens_micro$S
rownames(sens_micro$T) = colnames(sens_micro$X)
sens_micro$T

# notice how p12 is by far the most sensitive parameter

# graph the most sensitive parameter
tmp = cbind.data.frame(sens_micro$X, pop12=sens_micro$y)
ggplot(tmp, aes(p12, pop12))+geom_point()+labs(x="Survivability of 1 to older",y="pop after 12 months")
```


