---
title: "Sensitivity using Sobol"
output:
  slidy_presentation:
    highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
  beamer_presentation:
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(sensitivity)
library(tidyverse)
library(lhs)
library(purrr)
```

# Sobol

Sobol sensitivity analysis is similar to LHS approach as a way to efficiently sample parameter space 

Sobol is more general - it can use measured parameter samples (we will approximate sampling here by using known distributions but you could use actual samples!)

Sobol is a *variance-based* method

Sobol quantifies sensitivity by 
breaking the variance of output into 

# Sobol Indices - Quantifying Sensitivity

Indices are computed using parameter-output variance relationships; they are estimates so the indices themselves
have uncertainty bounds

Several Indices

<span style="color: green;">First order sensitivity or Main Effect</span>

* variance associated directly with parameter alone 
* fraction associated with each parameter, sum to 1 (although because it is estimated can be slightly more or less)

<span style="color: green;">Total Effect</span>

* variance associated with parameter and interaction with other parameters 
* sum can be more than 1 if parameters interact

<span style="color: green;">Second Order Indices</span>

* less used but quantify how parameter pair-wise parameter interactions contribute to output variation




# Sobol in R

* *Sensitivity* package

* Sobol Indices require estimation - and there are different methods to do that

* The *Sensitivity* package has several of those

* today we will use *sobolSalt* - which uses a method by Saltelli (who has written extensively on Sensitivity analysis)

* R help pages for *Sensitivity* provide many good references

* This is a nice overview paper

[Variance Based Methods](https://www.sciencedirect.com/science/article/pii/S0010465509003087?casa_token=xG4MJV4hwJgAAAAA:Tq_CLHTqtM1wDvhzviZFgm6sVBdOoYzQdFfOGplEi5OUxLSWxaKITce-CzxLxfS6993rNnbK1Q)

# Sobol - how to


* run Sobol to get parameter sets in a sensitivity analysis object
* run model with those parameter sets
* *tell* the senstivity object about results associated with each parameter set
* look at sensitivity analysis indices from Sobol

Generation of parameter sets slightly different

  * generate **two** samples of parameter sets by 
  samplng from apriori (expected) distributions
  * these would be the "hypothetical" distributions based on assumptions about the data
  * ideally these would be distributions that you actually sampled

#  Example 
Atmospheric Conductance as a function of windspeed, vegetation height and parameters

First lets get the parameters

```{r}
source("../R/Catm.R")

# generate two examples of random number from parameter distributions

np=1000
k_o = rnorm(mean=0.1,sd=0.1*0.1, n=np)
k_d = rnorm(mean=0.7, sd=0.7*0.1, n=np)
v = rnorm(mean=200, sd=10, n=np)
height = runif(min=5, max=15, n=np)


X1 = cbind.data.frame(k_o, k_d, v, height=height)

# repeat sampling
k_o = rnorm(mean=0.1,sd=0.1*0.1, n=np)
k_d = rnorm(mean=0.7, sd=0.7*0.1, n=np)
v = rnorm(mean=200, sd=10, n=np)
height = runif(min=5, max=15, n=np)

X2 = cbind.data.frame(k_o, k_d, v, height=height)

# there are different versions of sobol functions that have different approaches for estimating parameters and indices, we use an approach implemented by jansen 

sens_Catm_Sobol = sobolSalt(model = NULL, X1, X2, nboot = 100)


# Take a look at the Sobol generated sensitivity object
# your parameters sets for sensitivity analysis are in X
```

# Sobol Indices

now run model for Sobol generated parameter sets and compute indices

* pay attention to values of the indices and confidence intervals
  * if 0 is within the confidence interval, parameter uncertainty is not influencing output
  
* substantial differences between total effect and first order indices suggest parameter interactions

TIP: a useful plotting strategy is to plot model output against parameter with the highest total effect
and then use the parameter with second highest total effect for color

```{r sobolrun}
# run model for all parameter sets
# make sure you give the parameters names

parms = as.data.frame(sens_Catm_Sobol$X)
colnames(parms)= colnames(X1)
res = pmap_dbl(parms, Catm)


sens_Catm_Sobol = sensitivity::tell(sens_Catm_Sobol,res, res.names="ga")

# main effect:  partitions variance (main effect without co-variance) - sums approximately to one
sens_Catm_Sobol$S
# useful to add names
row.names(sens_Catm_Sobol$S) = colnames(parms)
sens_Catm_Sobol$S

# total effect - accounts for parameter interactions
row.names(sens_Catm_Sobol$T) = colnames(parms)
sens_Catm_Sobol$T

# Both the main effect and total effect can tell us something about how the parameter influences results


print(sens_Catm_Sobol)

 
# graph two most sensitive parameters
both = cbind.data.frame(parms, gs=sens_Catm_Sobol$y)

# look at response of conductance to the two most important variables
ggplot(both, aes(v,gs, col=height))+geom_point()+labs(y="Conductance (mm/s)", x="Windspeed")

```

# Some example articles
```{r}

```

 [Snow modeling](Snow_sobol.pdf)
 
 [Building Cooling Energy Mdoel](building_energy_sobol.pdf)
 
 [Uranium in Groundwater Model](Uranium_water_sobol.pdf)
 
 

# Second order indices

Optional for this course

If you want to also compute a second order indices you need to use a different variation. (scheme=B)

There are multiple implementation of Sobol in the *senstivity* package, they can be more or less stable
I find *sobolSalt* works well

```{r, alternative}

sens_Catm_Sobol2 = sobolSalt(model = NULL, X1, X2, nboot = 100, scheme="B" )

parms = as.data.frame(sens_Catm_Sobol2$X)
colnames(parms)= colnames(X1)
res = pmap_dbl(parms, Catm)


sens_Catm_Sobol2 = sensitivity::tell(sens_Catm_Sobol2,res, res.names="ga")

# main effect:  partitions variance (main effect without co-variance) - sums approximately to one
row.names(sens_Catm_Sobol2$S) = colnames(parms)
sens_Catm_Sobol2$S

# total effect - accounts for parameter interactions
row.names(sens_Catm_Sobol2$T) = colnames(parms)
sens_Catm_Sobol2$T

# second order parameters interaction in controlling sensitivity
# parameters are in order, interactiosn are small here
sens_Catm_Sobol2$S2

```
