---
title: 'Post Calibration Workshop'
output:
  slidy_presentation:
    highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
  beamer_presentation:
    highlight: pygments
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(sensitivity)
library(tidyverse)
library(lubridate)
library(reldist)
library(purrr)
library(ggpubr)
```

## Recall our investigation of calibration

I provided you with streamflow estimates from a hydrologic model, run many times with different parameters wets

Your testk:

* we defined ranges for parameters
* we defined measures of fit
* calculated "fit" for each parameter set
* chose the best parameter set

Essentially optimization


#  Original calibration script


I computed low flow metrics and based my calibration selection on how well the model
matched several low flow measures of fit - all combined into a single measure


```{r multiple}
source("../R/nse.R")

source("../R/relerr.R")

source("../R/cper.R")
source("../R/compute_lowflowmetrics_all.R")

source("../R/get_waterYearDay .R")

# observed data set
sager = read.table("../Data/sager.txt", header=T)
head(sager)

# add date
sager = sager %>% mutate(date = paste(day,month,year, sep="/"))
sager$date = as.Date(sager$date,"%d/%m/%Y")

# multiple results - lets say we've run the model for multiple years, 
#each column  is streamflow for a different parameter set
msage = read.table("../Data/sagerm.txt", header=T)

# keep track of number of simulations (e.g results for each parameter set) 
# use as a column names
nsim = ncol(msage)
snames = sprintf("S%d",seq(from=1, to=nsim))
colnames(msage)=snames


# lets say we know the start date from our earlier output
msage$date = sager$date
msage$month = sager$month
msage$year = sager$year
msage$day = sager$day
msage$wy = sager$wy

# lets add observed
msage = left_join(msage, sager[,c("obs","date")], by=c("date"))

head(msage)

# how can we plot all results - lets plot water year 1970 otherwise its hard to see
msagel = msage %>% pivot_longer(cols=!c(date, month, year, day,wy), names_to="run", values_to="flow")
# subset for split sample calibration
short_msage = subset(msage, wy < 1975)

# compute performance measures for output from all parameters

res = short_msage %>% select(-date, -month, -day, -year, -wy, -obs ) %>%
  map_df(compute_lowflowmetrics_all, o=short_msage$obs, month=short_msage$month, day=short_msage$day, year=short_msage$year, wy=short_msage$wy)

# note here we use map_df to get a dataframe back 


# interesting to look at range of metrics - could use this to decide on
# acceptable values
summary(res)
# we can add a row that links with simulation number
res$sim = snames

# graph range of performance measures
resl = res %>% pivot_longer(-sim, names_to="metric", values_to="value")

ggplot(resl, aes(metric, value))+geom_boxplot()+facet_wrap(~metric, scales="free")


# select the best one based on the combined metric
best = res[which.max(res$combined),]

# running the model forward
# so we can look at the full time series

# lets start with streamflow estimates from best performing parameter set
 ggplot(msage, aes(date, msage[,best$sim])) + geom_line()+geom_line(aes(date, obs), col="red") 

 
# for comparison lets consider how worst and best parameters perform for subsequent simulations
# focusing specifically on August streamflow
 worst = res[which.min(res$combined),]
 
 compruns = msage %>% select(best$sim, worst$sim, date, obs, month, day, year, wy)
 compruns = subset(compruns, wy > 1970)
 compruns_mwy = compruns %>% select(-c(day,date, year)) %>% group_by(month, wy) %>% summarize(across(everything(), mean))
 
 compruns_mwyl = compruns_mwy %>% pivot_longer(cols=!c(month,wy), names_to="sim", values_to="flow")
 compruns_mwyl %>% subset(month==8) %>% ggplot(aes(sim,flow ))+geom_boxplot()+labs(ylab="August flow")
 




```

# Look at class selected parameters

```{r, classmthy}

tmp = c("S6, S29, S32, S10, S95, S78, S89, S82, S1, S2, S95, S21, S99, S87, S97")
class_good_sim = unlist(strsplit(tmp, "[ ,]+"))
tmp = c("S29 S78 S32")
 class_really_good_sim =unlist(strsplit(tmp, "[ ,]+"))
 


 compruns_mwy = msage %>% select(-c(day,date, year)) %>% group_by(month, wy) %>% summarize(across(everything(), mean))
 
 compruns_mwyl = compruns_mwy %>% pivot_longer(cols=!c(month,wy), names_to="sim", values_to="flow")
 compruns_mwyl %>% subset(month==8) %>% ggplot(aes(sim,flow ))+geom_boxplot()
 
compruns_mwyl = compruns_mwyl %>% mutate(mgood=ifelse(sim %in% class_really_good_sim, TRUE, FALSE))
compruns_mwyl = compruns_mwyl %>% mutate(good=ifelse(sim %in% class_good_sim, TRUE, FALSE))
  
compruns_mwyl %>% subset(month==8) %>% ggplot(aes(sim,flow, fill=interaction(good,mgood) ))+geom_boxplot()
 
compruns_mwyl %>% subset((month==8) & (good | sim=="obs")) %>% ggplot(aes(sim,flow, fill=interaction(good,mgood) ))+geom_boxplot() + scale_fill_discrete(labels=c("Observed","Good","Multi-Good"))+labs(fill="Calibration Set", y="August Flow")

compruns_mwyl %>% subset((month==1) & (good | sim=="obs")) %>% ggplot(aes(sim,flow, fill=interaction(good,mgood) ))+geom_boxplot() + scale_fill_discrete(labels=c("Observed","Good","Multi-Good"))+labs(fill="Calibration Set", y="January Flow")

all_runs = subset(compruns_mwyl, sim != "obs") %>% group_by(month,wy) %>% summarize(flow=mean(flow),good=FALSE, mgood=FALSE)
unselected_runs = subset(compruns_mwyl, sim != "obs" & !good) %>% group_by(month,wy) %>% summarize(flow=mean(flow),good=FALSE, mgood=FALSE)
selected_runs = subset(compruns_mwyl, sim != "obs" & good) %>% group_by(month,wy) %>% summarize(flow=mean(flow),good=FALSE, mgood=FALSE)
all_runs$sim = "all"
unselected_runs$sim = "unselected"
selected_runs$sim = "selected"
tmp = rbind.data.frame(all_runs, unselected_runs, subset(compruns_mwyl, sim==best$sim | sim=="obs"), selected_runs)
tmp %>% subset((month %in% c(8))) %>% ggplot(aes(sim,flow, fill=sim ))+geom_boxplot()+ scale_fill_discrete(labels=c("Unselected","Good","Multi-Good"))+labs(fill="Calibration Set", y="August Flow")



 
 
```

# look at water year day

```{r classwyd}
msage$wyd = get_waterYearDay(msage$date, wyMonth=10)

 compruns_wyd = msage %>% select(-c(day,date, year, month, wy)) %>% group_by(wyd) %>% summarize(across(everything(), mean))
 
 compruns_wydl = compruns_wyd %>% pivot_longer(cols=!c(wyd), names_to="sim", values_to="flow") 
compruns_wydl = compruns_wydl %>% mutate(mgood=ifelse(sim %in% class_really_good_sim, TRUE, FALSE))
compruns_wydl = compruns_wydl %>% mutate(good=ifelse(sim %in% class_good_sim, TRUE, FALSE))
all_runs = subset(compruns_wydl, sim != "obs") %>% group_by(wyd) %>% summarize(flow=mean(flow),good=FALSE, mgood=FALSE)
unselected_runs = subset(compruns_wydl, sim != "obs" & !good) %>% group_by(wyd) %>% summarize(flow=mean(flow),good=FALSE, mgood=FALSE)
all_runs$sim = "all"
unselected_runs$sim = "unselected"

selected_runs = subset(compruns_wydl, sim != "obs" & good) %>% group_by(wyd) %>% summarize(flow=mean(flow),good=FALSE, mgood=FALSE)
selected_runs$sim = "allselected"

tmp = rbind.data.frame(compruns_wydl, all_runs[,colnames(compruns_wydl)], 
                       unselected_runs[,colnames(compruns_wydl)],
                        selected_runs[,colnames(compruns_wydl)],
                       subset(compruns_wydl, sim=="obs"))
tmp %>% subset( (good | (sim=="obs" | sim=="all" | sim=="unselected"))) %>% ggplot(aes(wyd,flow, color=sim) )+geom_line() 


tmp %>% subset( (sim=="allselected" | sim=="obs" | sim=="all" | sim=="unselected")) %>% ggplot(aes(wyd,flow, color=sim) )+geom_line() + labs(y="Mean Daily Flow (mm/day)")



```

# What we learn

* calibration results depend on the metric

* as a results, calibration can be imprecise and may depend no only on metric, but time available
for calibration

* keeping multiple parameters that give "reasonable" results provides more robust modeling results
