---
title: "Dynamicmodel2logistic"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(deSolve)

```

# Integration

What if we wanted to look at population in 20 years given an initial condition

Two options

* explicit solution to differential equation is known; e.g. you can integrate both sides of the equation! Not always possible but lets look at a case where it is possible

* must be solved by iteration; this is what we do when we can't integrate both sides 

# Explicit Solution is available

```{r explicit}
source("../R/exppop.R")

# gives population after any time given an initial population

# 20 rabbits, growth rate of 0.01 how many in 30 years
exppop(T=30, P0=20, r=0.01, K=1000)

# if we want to see how population evolves over time - generate a time series by running our model for each point in time

initialrabbits = 20
years = seq(from=1, to=100, by=2)
Ptime = years %>% map_dbl(~exppop( P0=initialrabbits, r=0.01, K=1000, T=.x))

# keep track of what times we ran
Ptime = data.frame(P=Ptime, years=years)

ggplot(Ptime, aes(years,P))+geom_point()+labs(x="years",y="Rabbit Population")

# try generating results for maximum and minimum possible r values to compare (guess at what you think)


max_r = 0.1
min_r = 0.01
K = 1000

tmp = years  %>% map_dbl(~exppop(r=max_r, P0=initialrabbits, K=K, T=.x))
Ptime$Pmaxr = tmp
tmp = years %>% map_dbl(~exppop(r=min_r, P0=initialrabbits, K=K, T=.x))
Ptime$Pminr = tmp

head(Ptime)
Ptimep = Ptime %>% gather(key="r",value="P",-years)
ggplot(Ptimep, aes(years,P, col=r))+geom_point()+labs(x="years",y="Rabbit Population")

# notice how populatin becomes unstable for high growth rates! 


```

Using a solver....when you can't do the integration by hand :)

For example, if you added a carrying capacity!

We can still do the integration by iteration - approximates it
Implement the differential equation as a function that

*returns the derivative (as a list)

*inputs time, the variable(s) and a parameter list

(it needs time even though you don’t use it)

I name them starting with *d* to remind myself that they
are computing a derivative


R has a solver called *ODE* for solving ordinary differential equations
frome package **desolve**

*ODE* requires

* initial conditions
* values of independent where we want values of dependent variable (e.g times where we want population)
* the derivative as a function
* a list that contains alll  parameter values (or if you have only one parameter then you can use a single value)

```{r usingode}


source("../R/dexppop.R")
library(deSolve)
initialrabbits = 20
years = seq(from=1, to=100, by=2)

# run the solver
Ptime = ode(y=initialrabbits, times=years, func=dexppop,parms=c(0.01))
head(Ptime)
colnames(Ptime)=c("year","P")

# notice that there are additional pieces of information year, including the method used for integration
attributes(Ptime)

# this also means you need to extract just the data frame for plotting
ggplot(as.data.frame(Ptime), aes(year,P))+geom_point()+labs(y="Population", "years")

# this also works (of course function can be by order)
Ptime=ode(initialrabbits, years, dexppop,0.03)
colnames(Ptime)=c("year","P")
ggplot(as.data.frame(Ptime), aes(year,P))+geom_point()+labs(y="Population", "years")

# how would we add carrying capacity here

```


You can play a bit with changing your function to something that you can't integrate "by hand"

BUT we might want more parameters

to work with ODE,
parmeters must all be input as a single list; similar to how we return multiple outputs from a function

see example below..lets add a carrying capacity


```{r}



source("../R/dexppop_play.R")

# create parameter list
initalrabbits=2
newparms = list(r=0.03, carry_capacity=300)

#apply solver
results=ode(initialrabbits, years, dexppop_play,newparms)
head(results)

# add more meaningful names
colnames(results)=c("year","P")

# plot
ggplot(as.data.frame(results), aes(year,P))+geom_point()+labs(y="Population", "years")


# try again with different parameters
alternativeparms = list(r=0.04, carry_capacity=500)
results2=ode(initialrabbits, years, dexppop_play,alternativeparms)


# look at results
head(results2)
colnames(results2)=c("year","P_parm2")

# plot
ggplot(as.data.frame(results2), aes(year,P_parm2))+geom_point()+labs(y="Population", "years")

# compare by combining into a single data frame
both = inner_join(as.data.frame(results), as.data.frame(results2))

both_p = both %>% gather(key="model", value="P", -year)
ggplot(both_p, aes(year,P, col=model))+geom_point()+labs(y="Population", "years")

# try playing on your own - modiffy the function in some way
```


## Difference Equations

Population models can be discrete (rather than continuous)
So we could implement them as difference equations and iterate


```{r iteratie}
source("../R/discrete_logistic_pop.R")
# notice how a for loop is used to iterate

# how many rabbits after 50 years given a growth of 0.1
# starting with 1 rabbit - but a carrying capcity of 500

discrete_logistic_pop

discrete_logistic_pop(P0=1, r=0.05, K=200, T=50)

# save results
discrete_result = discrete_logistic_pop(P0=1, r=0.05, K=200, T=50)

# lets also keep the parameters for use later
P0=1
r=0.05
K=200
T=50

```

## Differential Equation comparison

Remember we have 3 ways now to calculate population

analytical solution - based on integration (exppop.R) BEST

using an ode solver for numerical approximation (exppop_play.R)

numerical integration using in discrete steps (discrete_logistic_pop.R)
```{r closed}

source("../R/exppop.R")

exppop(P0=P0, r=r, K=K, T=T)
analytic_result = exppop(P0=P0, r=r, K=K, T=T)

analytic_result
discrete_result

# why are they different
# look at trajectories

growth_result = data.frame(time=seq(from=1,to=100))

growth_result$Panalytic = growth_result$time %>% map_dbl(~exppop( P0=1,r=0.05, K=200,T=.x ))

growth_result$Pdiscrete = growth_result$time %>% map_dbl(~discrete_logistic_pop( P0=1,r=0.05, K=200,T=.x ))

tmp = growth_result %>% gather(key="Ptype",value="P",-time)
ggplot(tmp, aes(time,P, col=Ptype))+geom_point()

# try running them for longer time periods to see what happens 
# change the value of r, K , P0 - see how it effects the results
```

Finally look at continuous derivative using ODE solve
Needs
*initial condtions
*differential equation
*parameters

```{r}

source("../R/dexppop_play.R")

dexppop_play
# set up using the same parameters
pcompare = list(r=r,carry_capacity=K)


# now run our ODE solver
result = ode(P0, growth_result$time, dexppop_play, pcompare)
head(result)

# we already have time - so just extract population
growth_result$Pdifferential = result[,2]

# comapre all 3 approaches
tmp = growth_result %>% gather(key="Ptype",value="P",-time)
ggplot(tmp, aes(time,P, col=Ptype))+geom_point()

# notice Pdifferential is closer to Pa
```

All differential and difference equations are approximations
The analytical solution is exact

Notice that differential equations is a bit more accurate!

## Diffusion

Diffusion can be implemented as a partial differential equation 
Complicated to solve - but there are tool in R for specific types of partial differential equations
![Reactive Transport Example](https://cran.r-project.org/web/packages/ReacTran/index.html)

More info on differential equations in R
![online book differential equation in R](http://www-labs.iro.umontreal.ca/~mignotte/IFT2425/Documents/Solving_Differential_Equations_In_R_Soetaert_K.pdf)

But we can appoximate this with a difference equation - and iterative to get an estimate of how diffusion works (see class lecture pdf)


Now for partial derivatives - time and space! it gets much more tricky ...beyond this course 

Example of Diffusion - difference equation implementation to see what some issues can be


```{r}

source("../R/diffusion.R")

# run our diffusion model (iterative difference equation) with initial concentration of 10, for 8 timestep (size 1m), and 10 space steps (size 1s)
# using diffusion parameters 0.5 s/m2, 10 m2
result = diff1(initialC=10, nx=10, dx=1, nt=8, dt=1, D=0.5, area=10)

# a list is returned with our 3 data frames for concentration (conc), qin and qout
result

# used filled contour to plot results
head(result$conc)
filled.contour(result$conc, xlab="Time", ylab="Distance")

# or if you prefer this orientation (Distance on x axis)
filled.contour(t(result$conc), ylab="Time", xlab="Distance")

# changes diffusivity and other parameters particularly
# diffusivity, dx and dt
res=diff1(initialC=10,nx=10,dx=1,nt=10,dt=30,D=0.006,area=1)

filled.contour(res$conc, xlab="Time", ylab="Distance")


# we can also see how much material moved from place to place each time step
filled.contour(res$qin, xlab="Time", ylab="Distance")

# what if we increase diffusivity
resfast=diff1(initialC=10,nx=10,dx=0.5,nt=10,dt=10,D=0.08,area=1)
filled.contour(resfast$conc, xlab="Time", ylab="Distance")


# play with time step, space step and parameters


```

```{r}

# Discretization Issue Example
resunstable=diff1(initialC=10,nx=10,dx=1,nt=10,dt=10,D=0.8,area=1)
filled.contour(resunstable$conc, xlab="Time (fraction of hour)",ylab="Distance Along Path (m)", main="Pollutant Diffusion")

# this illustrates the problem with difference equations (and the challenges that methods for numerical integration try to overcome)
# if things are changing quickly we need to use much smaller time, space steps to avoid overshoot and instability

# so lets cut our step size by 10 (dt) (but then add 10 more steps (nx) to cover the same distance)
resunstable=diff1(initialC=10,nx=100,dx=1,nt=10,dt=1,D=0.8,area=1)
filled.contour(resunstable$conc, xlab="time",ylab="Distance Along Path")

```

