---
title: "Sensitivity with ODEs"
output:
  slidy_presentation:
    highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
  beamer_presentation:
    highlight: pygments
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(deSolve)
library(sensitivity)
```

# Sensitivity Analysis of a differential equation 

We can apply sensitivity analysis to a differential equation

A key issue where is sensitivity of what?

Dynamic models often give you many many outputs - time series (streamflow every day for a year, popultation for 30 years) - or output over space (concentration every 10 meters)

So if we are asking 'sensitivity of what' we need to summarize results in some way (reduce their dimensionality )

Ideas?


# Some options for reducing output dimensionality (summarizing output)

Depends on what is important for your model application

* max

* mean

* min

* total

* variation

* time it takes for something to happen

So a key step in sensitivity analysis with a dynamics model is summarizing results into a few key measures

Its useful to turn that summarizing workflow into a function


## Workflow

* obtain parameter sets (from sobel of LHS)

* build a function that will extract the information (metrics) you want from your dynamic model (output of the ode)

* create a data structure to store the metrics for each parameter set - lets call it metrics (but could be anything)

* run ODE for each parameter sets to fill in this metrics data structure

* send the metrics data structure back to the sensitivity analysis object (from sobel or LHS)

* plot and analyze results

# Example with our population ODE

```{r sen}
source("../R/dexppop_play.R")

dexppop_play

# lets start with sobel 
library(sensitivity)

# come up with first set of sample parameters
# we will assume that we know the initial population,

Pinitial=10

# want to learn about sensitivity to growth rate (r) and carrying capacity 
# set the number of parameters
np=100
carry_capacity = rnorm(mean=400, sd=50, n=np)
r = rnorm(mean=0.01, sd=0.005, n=np)
X1 = cbind.data.frame(r=r, carry_capacity=carry_capacity)

# repeat to get our second set of samples
carry_capacity = rnorm(mean=400, sd=50, n=np)
r = rnorm(mean=0.02, sd=0.005, n=np)
X2 = cbind.data.frame(r=r, carry_capacity=carry_capacity)

# create our sobel object and get sets ofparameters for running the model

sens_P = sobolSalt(model = NULL,X1, X2, nboot = 300)

# our parameter sets are
head(sens_P$X)

# lets add names 
colnames(sens_P$X) = c("r","carry_capacity")

# run our differential equation and keep the output
# BUT
# what output do we want  to keep
# how about maximum population if we run the model for 200 years, and how many years to get to the carrying capacity

# for illustration lets look at running just one parameter sets and summarizing results
sens_P$X[1,]
# recall ODE needs ALL of our parameters in a single list 
# initial population and timnes for which we want output 
Pinitial

# gets results for 200 years (evaluating every year)
simtimes = seq(from=1, to=200)
parms = list(r=sens_P$X[1,"r"], carry_capacity=sens_P$X[1,"carry_capacity"])

# or
parms = list(r=as.data.frame(sens_P$X)$r[1], carry_capacity=as.data.frame(sens_P$X)$carry_capacity[1])

result = ode(y=Pinitial, times=simtimes, func=dexppop_play, parms=parms)

head(result)
colnames(result)=c("time","P")
# turn it into a data frame
result = as.data.frame(result)
ggplot(result, aes(time, P))+geom_point()

# extra our metrics of interest  from this
# maximum population it gets to
maxpop = max(result$P)
maxpop

# years required to get to the maximum poputation
# which.max will tell us when max population occur
idx = which.max(result$P)
# turn this index into a year (might be the same if time step in 1 but just in case it isn't)
maxyear = result$time[idx]
maxyear
```

# Compute our metric for all the parameter sets

What if we want to run for all parameters

Lets create two additional functions that will help us

* a function that computes the metrics we want

* a function that runs our ode solver and computes the metrics (I call it a wrapper function as it is really just a workflow/wrapper to call ode solver and then compute metrics)
```{r sen2}
# turn computing our metrics into a function

compute_metrics = function(result) {
  maxpop = max(result$P)
idx = which.max(result$P)
maxyear = result$time[idx]
return(list(maxpop=maxpop, maxyear=maxyear))}

# try it on our first parameter set
compute_metrics(result)

# great but we need to apply the ode and this function for all of our parameters



# define a wrapper function to do everything we need - run solver and compute metrics - and send back results for each parameter

p_wrapper = function(r,carry_capacity, Pinitial, simtimes, func) {
    parms = list(r=r, carry_capacity=carry_capacity)
    result = ode(y=Pinitial, times=simtimes, func=func, parms=parms) 
    colnames(result)=c("time","P")
  # get metrics
  metrics=compute_metrics(as.data.frame(result))
  return(metrics)
}

# now use pmap as we did before

allresults = as.data.frame(sens_P$X) %>% pmap(p_wrapper, Pinitial=Pinitial, simtimes=simtimes, func=dexppop_play)

# extract out results from pmap into a data frame
allres = allresults %>% map_dfr(`[`,c("maxpop","maxyear"))


# create boxplots
tmp = allres %>% gather(key="metric", value="value")
ggplot(tmp, aes(metric, value, col=metric))+geom_boxplot()
```

Compute the sobol indicies for each metric

```{r sen3}
# sobol can only handle one output at a time  - so we will need to do them separately

sens_P_maxpop = sensitivity::tell(sens_P,allres$maxpop)

# first-order indices (main effect without co-variance)
sens_P_maxpop$S

# total sensitivity index -note that this partitions the output variance - so values sum to 1
sens_P_maxpop$T




# create another one for max year
sens_P_maxyear = sensitivity::tell(sens_P,allres$maxyear)
# first-order indices (main effect without co-variance)
sens_P_maxyear$S

# total sensitivity index -note that this partitions the output variance - so values sum to 1
sens_P_maxyear$T

```

# Error messages from ODE

*In lsoda(y, times, func, parms, ...) :
  an excessive amount of work (> maxsteps ) was done, but integration was not successful - increase maxsteps*
  
Suggest that the solver (numerical integration) had issues

  * increasing maxsteps can help
  
        *result = ode(y=Pinitial, times=simtimes, func=func, parms=parms, maxsteps=100000) *
        
  * trying different methods
  
        *result = ode(y=Pinitial, times=simtimes, func=func, parms=parms, method="daspk")*
        
  * "stiff" problems are harder for numerical integration to solve - (small changes have big impacts);
    our threshold carrying capacity does that - change to make is a more smooth transition
    
      thoughts on how to do this?
    
  
  
# For next class (not to hand in - just practice on your own)

Go through in-class examples; 

  * try varying parameter ranges, 
  
  * try changing the population growth function to smooth the transition to no growth when a carrying capacity is approached ; 
  
  * try developing a different metric
  
We will go over this in class on Tuesday


Want to go further?....

Try using Sobol for the diffusion model - what would be your metric?
