---
title: "More on Diffusion"
output:
  slidy_presentation:
    highlight: pygments
  html_document: default
  pdf_document: default
  ioslides_presentation:
    highlight: pygments
  beamer_presentation:
    highlight: pygments
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(deSolve)
library(sensitivity)
```

# Dynamics Models

* Diffusion example illustrates the challenge of numerical integration

* We see evidence of "overshoot" 

* Correct by reducing the time step (but then we have to increase the number of time steps to cover the same period)
  * recall total time is number of time steps (nt) multiplied by time interval (dt)
  

# Diffusion Example

```{r}
source("../R/diffusion.R")


# Change parameters (diffusivity D, and space and time steps (dx, dt))

res=diff1(initialC=100,nx=10,dx=1,nt=10,dt=30,D=0.001,area=1)
filled.contour(res$conc, xlab="Time", ylab="Distance", main = "Concentration through time and space")


# we can also see how much material is moving in to each cell at each time step
filled.contour(res$qin, xlab="Time", ylab="Distance", main="Qin - material coming in ")

# we can also see net amount of material moved from place to place each time step
filled.contour(res$qin-res$qout, xlab="Time", ylab="Distance", main="Qin-Qout NET ")

# what if we increase diffusivity
resfast=diff1(initialC=100,nx=10,dx=0.5,nt=10,dt=10,D=0.08,area=1)
filled.contour(resfast$conc, xlab="Time", ylab="Distance", main="Concentration through time and space")
filled.contour(resfast$qin, xlab="Time", ylab="Distance", main="Qin")


# this illustrates the problem with difference equations (and the challenges that methods for numerical integration try to overcome)
# if things are changing quickly we need to use much smaller time, space steps to avoid overshoot and instability

# so lets cut our step size by 10 (dt) (but then  multiply number of steps (nx) to cover the same distance)
resfast_fixtime=diff1(initialC=100,nx=10,dx=0.5,nt=100,dt=1,D=0.08,area=1)
filled.contour(resfast_fixtime$conc, xlab="time",ylab="Distance Along Path", main="Concentration through time and space")

filled.contour(resfast_fixtime$qin, xlab="Time", ylab="Distance", main="Qin")
filled.contour(resfast_fixtime$qin-resfast_fixtime$qout, xlab="Time", ylab="Distance", main="Net Transport")


```

# Extracting meaning from time series output

Useful to brainstorm about what is important


For example 

* time it takes to evenly diffuse?

How would we implement that?

# Extracting information from space-time results

* pictures can be hard to interpret

* summarizing over one of the dimensions (either space or time) can help

* looking at a single trajectory through time

* looking at spatial variation for one point in time

* looking at spatial variation for multiple points in time

# Try it


```{r one option}
#View(resfast_fixtime$conc)


# graph a single point in space through time
# single column (time)
plot(resfast_fixtime$conc[,3], ylab="Concentration for a location 2 spatial units from origin (1+2")

# plot all trajectories
# add a time column to concentration data frame and transform for plotting
resl = as.data.frame(resfast_fixtime$conc) %>% mutate(time=seq(from=1, to=100)) %>% pivot_longer(-time, names_to="distance", values_to="conc")
ggplot(resl, aes(time, conc, col=distance))+geom_line()

# plot all places at each point in time
ggplot(resl, aes(time, conc, group=time))+geom_boxplot()

# use apply to calculate the spatial variation for each row (e.g for each time point) 
cvar = resfast_fixtime$conc %>% apply(1,var)
cmean = resfast_fixtime$conc %>% apply(1, mean)

spatial_aver = cbind.data.frame(cvar, cmean, time=seq(from=1,to=100))
length(cvar)
# notice its the same as the number of time units (nt) used above

# plot spatial variation through time
ggplot(spatial_aver, aes(time, cvar))+geom_line()+labs(y="Spatial Variation")

# plot coefficient of variation (so standard deviation divided by the mean)
ggplot(spatial_aver, aes(time, 100*sqrt(cvar)/cmean))+geom_line()+labs(y="COV (as percent")


```

