---
title: "moredynamic"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# additional parameters 

You can play a bit with changing your function to something that you can't integrate "by hand"

BUT we might want more parameters

to work with ODE,
parameters must all be input as a single list; similar to how we return multiple outputs from a function

see example below..lets add a carrying capacity

# R code with carrying capacity

```{r}



source("../R/dexppop_play.R")

dexppop

# create parameter list
initalrabbits=2
newparms = list(r=0.03, carry_capacity=300)

#apply solver
results=ode(initialrabbits, years, dexppop_play,newparms)
head(results)

# add more meaningful names
colnames(results)=c("year","P")

# plot
ggplot(as.data.frame(results), aes(year,P))+geom_point()+labs(y="Population", "years")


# try again with different parameters
alternativeparms = list(r=0.04, carry_capacity=500)
results2=ode(initialrabbits, years, dexppop_play,alternativeparms)


# look at results
head(results2)
colnames(results2)=c("year","P_parm2")

# plot
ggplot(as.data.frame(results2), aes(year,P_parm2))+geom_point()+labs(y="Population", "years")

# compare by combining into a single data frame
both = inner_join(as.data.frame(results), as.data.frame(results2))

both_p = both %>% gather(key="model", value="P", -year)
ggplot(both_p, aes(year,P, col=model))+geom_point()+labs(y="Population", "years")

# try playing on your own - modify the function in some way
```

# Difference Equations 

What is ODE doing? (iterating in 'smart ways')

Similar to "difference equations"

## Difference Equations

Population models can be discrete (rather than continuous)
So we could implement them as difference equations and iterate


```{r iteratie}
source("../R/discrete_logistic_pop.R")
# notice how a for loop is used to iterate

# how many rabbits after 50 years given a growth of 0.1
# starting with 1 rabbit - but a carrying capcity of 500

discrete_logistic_pop

discrete_logistic_pop(P0=1, r=0.05, K=200, T=50)

# save results
discrete_result = discrete_logistic_pop(P0=1, r=0.05, K=200, T=50)

# lets also keep the parameters for use later
P0=1
r=0.05
K=200
T=50

```

# Differential Equation, Difference (Iteration by hand) comparison

Remember we have 3 ways now to calculate population

analytical solution - based on integration (exppop.R) BEST

using an ode solver for numerical approximation (exppop_play.R)

numerical integration using in discrete steps (discrete_logistic_pop.R)
```{r closed}

source("../R/exppop.R")

exppop(P0=P0, r=r, K=K, T=T)
analytic_result = exppop(P0=P0, r=r, K=K, T=T)

analytic_result
discrete_result

# why are they different
# look at trajectories

growth_result = data.frame(time=seq(from=1,to=100))

growth_result$Panalytic = growth_result$time %>% map_dbl(~exppop( P0=1,r=0.05, K=200,T=.x ))

growth_result$Pdiscrete = growth_result$time %>% map_dbl(~discrete_logistic_pop( P0=1,r=0.05, K=200,T=.x ))

tmp = growth_result %>% gather(key="Ptype",value="P",-time)
ggplot(tmp, aes(time,P, col=Ptype))+geom_point()

# try running them for longer time periods to see what happens 
# change the value of r, K , P0 - see how it effects the results
```

# Compare analytical, difference and ODE

Finally look at continuous derivative using ODE solve
Needs
*initial condtions
*differential equation
*parameters

```{r}

source("../R/dexppop_play.R")

dexppop_play
# set up using the same parameters
pcompare = list(r=r,carry_capacity=K)


# now run our ODE solver
result = ode(P0, growth_result$time, dexppop_play, pcompare)
head(result)

# we already have time - so just extract population
growth_result$Pdifferential = result[,2]

# comapre all 3 approaches
tmp = growth_result %>% gather(key="Ptype",value="P",-time)
ggplot(tmp, aes(time,P, col=Ptype))+geom_point()

# notice Pdifferential is closer to Panalytic than Pdiscrete

```