---
title: "harvest"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(sensitivity)
library(tidyverse)
library(deSolve)
```

Generate and run a model of harvest

```{r harvest}

# fixed amount per year
source("../R/dharvest.R")
dharvest

# try it out
tm = seq(from=1, to=500)
Pinitial = 1
gps = list(harv=0.18, K=100, r=0.2)
res = ode(Pinitial,tm, dharvest, gps)
colnames(res)=c("time","carbon")

ggplot(as.data.frame(res), aes(time, carbon))+geom_point()

# what if we harvest a a much greater rates

# lets vary the harvest rates from 0.0 to 0.4
harvestr = seq(from=0.0, to=0.4, by=0.025)

# save all of the trajectories


# use a wrapper function to just return the carbon trajectories
getcarbon = function(Pinitial, tm, harv, K, r, hfunc) {
  gps = list(harv=harv, K=K, r=r)
  res = ode(Pinitial,tm, hfunc, gps)
  colnames(res)=c("time","carbon")
  res=as.data.frame(res)
  return(carbon=res$carbon)
}

# apply this function to all harvest values
res = harvestr %>% map_dfc(~getcarbon( Pinitial=Pinitial, tm=tm, K=100, r=0.2, hfun=dharvest, harv=.x))
# rows are time, columns are carbon for each harvest scenario
colnames(res)=harvestr
res=as.data.frame(res)
res$time=tm
# put in to a form where we can plot
resl = gather(res, key="harvestr",value="carbon", -time)
ggplot(resl,aes(time, carbon, col=harvestr))+geom_line()


# notice that stable forest value changes with harvest rates

# notes that some forests are not stable - (or stablity is zero)

# see this at the beginning - plot the first 10 years
ggplot(subset(resl, time < 10),aes(time, carbon, col=harvestr))+geom_line()

```

Try different harvest rates...notice how carbon growth changes...what does stability mean - can you find parameter sets that lead to a stable non-zero forest

We could do some sensitivity analysis to see how harvest rate, and growth rate interact to control where the forest ends up after 10 years and 50 years (short and long planning horizons)

We will use our compute metrics and wrapper to make this easy

```{r harvestsend}

# fixed amount per year
source("../R/dharvest.R")


# lets assume a uniform distribution of harvest rates
# and of normal growth rates
np=200

r = rnorm(mean=0.3, sd=0.05, n=np)
harv = runif(min=0.0, max=0.4, n=np)
K = 
X1 = cbind.data.frame(r=r, harv=harv)

# repeat to get our second set of samples
r = rnorm(mean=0.3, sd=0.05, n=np)
harv = runif(min=0.0, max=0.4, n=np)
X2 = cbind.data.frame(r=r, harv=harv)

# create our sobel object and get sets ofparameters for running the model

sens_forest = sobolSalt(model = NULL,X1, X2, nboot = 300)
colnames(sens_forest$X)= c("r","harv")
# do a quick test
# try it out
tm = seq(from=1, to=50)
Pinitial = 1
parms = list(r=sens_forest$X[1,1], harv=sens_forest$X[1,2], K=100)
res = ode(y=Pinitial,times=tm, func=dharvest, parms=parms)
res=as.data.frame(res)
colnames(res)=c("time","C")

# compute our two metrics of interest - harvest after 10 and 50 years
compute_metrics = function(res) {
 C50 = res[50]
 C10 = res[10]
return(list(C50=C50, C10=C10))}

# use a wrapper function to just return the carbon trajectories
p_wrapper = function(r,harv, K, Pinitial, simtimes, func) {
    parms = list(r=r, K=K, harv=harv)
    result = ode(y=Pinitial, times=simtimes, func=func, parms=parms) 
    result=as.data.frame(result)
   colnames(result)=c("time","C")
  # get metrics
  metrics=compute_metrics(result$C)
  return(metrics)
}

# notice how we added in K, a parameter that we are NOT varying

# try it out
tm = seq(from=1, to=50)
Pinitial = 1


allresults = as.data.frame(sens_forest$X) %>% pmap(p_wrapper, K=100, Pinitial=Pinitial, simtimes=tm, func=dharvest)

# extract out results from pmap into a data frame
allres = allresults %>% map_dfr(`[`,c("C10","C50"))

tmp = allres %>% gather(key="metric", value="value")
ggplot(tmp, aes(metric, value, col=metric))+geom_boxplot()

# link with parameters to see how parameters together
# impact something we care aboue C50

allresp = cbind.data.frame(sens_forest$X, allres)
ggplot(allresp, aes(harv, C50, col=r))+geom_point()

# notice how we can see the impact of harvesting - and how growth rates reduce the sustainable harvest
```
