---
title: "predprey
"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
library(ggpubr)
library(tidyverse)
library(sensitivity)
library(lhs)
library(purrr)
library(ggpubr)
```

Predictor Prey - 


```{r}
source("../R/lotvmod.R")

# initial conditions
currpop = c(prey=10, pred=1)

# time points to see results
days = seq(from=1, to=100, by=1)

# set parameters
pars = c(rprey=0.5, alpha=0.3, eff=0.2,pmort=0.2, K=100)

# run the model
res = ode(func=lotvmod, y=currpop, times=days, parms=pars)

# graph the results
head(res)
# rearrange for easy plotting
ressimple = as.data.frame(res) %>% gather(key="animal", value="pop",-time)
p1=ggplot(ressimple, aes(time, pop, col=animal))+geom_line()

# why is it cyclical

# another way to look at results - relationship between populations

p2=ggplot(as.data.frame(res), aes(pred, prey))+geom_point()

# try other parameters to bring relative size of preditors higher


ggarrange(p1,p2)

```


Some more illustrations 

Preditor and Prey with Carrying Capacity


```{r examples}

source("../R/lotvmodK.R")

# initial conditions
currpop=c(prey=1, pred=1)

# set parameter list
pars = c(rprey=0.1, alpha=0.6, eff=0.8,pmort=0.4, K=20)

# times when you want to evaluate
days = seq(from=1,to=500)

# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars)

# estract the results
ressimple = as.data.frame(res) %>% gather(key="animal", value="pop",-time)

# graph both populations over time
p1=ggplot(ressimple, aes(time, pop, col=animal))+geom_line()
p1

# also look at relationships between preditor and prey population and use color for time 
# I will remove the legend here to make it easier to see 
p2 = ggplot(as.data.frame(res), aes(pred, prey, col=as.factor(round(time/10))))+geom_point()+theme(legend.position = "none")
p2

ggarrange(p1,p2)

# try with different parameter sets, can you create one where populations are stable - less cycling?

```


# Sensitivity analysis

Consider pred-prey 
BUT what will be the output - if we want to 'quantify sensitivity'
useful to look at a single value or set of value

for example
Max Prey Pop
Min Prey Pop


Remember steps

* Generate parameters (LHS, Sobel)
* Metrics function
* Wrapper Function
* Run wrapper function to get metrics for all paramater sets
* Graph and compute sensitivity statistics

```{r odesen}
source("../R/lotvmodK.R")
library(lhs)
library(epiR)

# create parameter sets...
factors = c("rprey","K","alpha","eff","pmort")
nsets=500

# create a latin hyper cube
sens_pp = randomLHS(nsets, length(factors))
colnames(sens_pp)=factors

# refine using sampling distributions for each parameter
sens_pp[,"rprey"]= qunif(sens_pp[,"rprey"], min=0.01, max=0.3)
sens_pp[,"K"]= qunif(sens_pp[,"K"], min=10, max=200)
sens_pp[,"alpha"]= qunif(sens_pp[,"alpha"], min=0.1, max=0.4)
sens_pp[,"eff"]= qnorm(sens_pp[,"eff"], mean=0.3, sd=0.01)
sens_pp[,"pmort"]= qunif(sens_pp[,"pmort"], min=0.05, max=0.45)

# lets create a metric and wrapper function as we did for our population models

# first our metrics 
# lets say we  want the maximum and minimum  of both predictor and prey

compute_metrics = function(result) {
  maxprey = max(result$prey)
  maxpred = max(result$pred)
  minprey = min(result$prey)
  minpred = min(result$pred)
return(list(maxprey=maxprey, minprey=minprey, maxpred=maxpred, minpred=minpred))}

# build a wrapper function


p_wrapper = function(rprey,alpha, eff, pmort, K, currpop, days, func) {
    parms = list(rprey=rprey, alpha=alpha, eff=eff, pmort=pmort, K=K)
    result = ode(y=currpop, times=days, func=func, parms=parms) 
    colnames(result)=c("time","prey","pred")
  # get metrics
  metrics=compute_metrics(as.data.frame(result))
  return(metrics)
}


# run our model for all parameters and extract the results
currpop=c(prey=1, pred=1)
days = seq(from=1,to=500)
allresults = as.data.frame(sens_pp) %>% pmap(p_wrapper, currpop=currpop, days=days, func=lotvmodK)
 
# take results back to unlisted form
allres = allresults %>% map_dfr(`[`,c("maxprey","minprey","maxpred","minpred"))


# range of response across parameter uncertainty
allresl = allres %>% gather(key="metric",value="pop")
ggplot(allresl, aes(metric, pop))+geom_boxplot()

# dealing with different scales
ggplot(allresl, aes(metric, pop, col=metric))+geom_boxplot()+facet_wrap(~metric, scales="free")
# plot cummulative densities

ggplot(allresl, aes(pop, col=metric))+stat_ecdf(geom="line")+facet_wrap(~metric, scales="free")

# compute PRCCs using epi.prcc
# lets do first for maxpred

epi.prcc(cbind.data.frame(sens_pp, allres$maxpred))

# try minprey
epi.prcc(cbind.data.frame(sens_pp, allres$minprey))


```



How do we think about stablity?

Populations don't change when derivatives are zero!

What conditions lead to BOTH derivatives being zero

For lotvmod
* dprey = rprey*prey -  alpha*prey*pred
* dpred = eff*alpha*prey*pred - pmort*pred

Make dprey and dpred equal to 0 and rearrange
* pred_i = rprey/alpha
* prey_i = pmort/(eff * alpha)

For lotvmodK

* dprey = rprey*(1-prey/K)*prey -  alpha*prey*pred
* dpred = eff*alpha*prey*pred - pmort*pred

Make dprey and dpred equal to 0 and rearrange

* prey_i = pmort/(eff*alpha)
* pred_i = rprey/alpha * (1-prey_i/K)
Try setting you initial conditions close to these values and see what happens


```{r stabilitynew}

# set parameter list
pars = data.frame(rprey=0.1, alpha=0.6, eff=0.8,pmort=0.4, K=200)

# now lets try initial conditions that will be stable
preyi = with(pars, pmort/(eff*alpha)) 
predi = with(pars, rprey/alpha*(1-preyi/K)) 
# times when you want to evaluate
days = seq(from=1,to=500)

# lets first see what happens when we start with 1 of each
currpop=c(prey=1, pred=1)
# run our differential equation solver
res = ode(func=lotvmodK, y=currpop, times=days, parms=pars)
# extract the results
res_smallstart = as.data.frame(res) %>% gather(key="animal", value="pop",-time)
# graph both populations over time
p1=ggplot(res_smallstart, aes(time, pop, col=animal))+geom_line()
p1

# lets first see what happens when we start our estimates of stable populations
stablepop = c(prey=preyi, pred=predi)
res = ode(func=lotvmodK, y=stablepop, times=days, parms=pars)
# estract the results
res_stablestart = as.data.frame(res) %>% gather(key="animal", value="pop",-time)
# graph both populations over time
p2=ggplot(res_stablestart, aes(time, pop, col=animal))+geom_line()
p2

# of course in this case these are not very realistic populations, so it is unlikely that you would get to something stable given parameters - but if they were different you might find a stable and realistic population

# try
```
And just for fun

Lets look at a Lorenz System with its interesting dynamics
```{r lorenze}
# lorenze
source("../R/lorenz.R")
pars = list(a=10,b=28,c=8/3)
res = ode(func=lorenz, c(x=0.1,y=0,z=0), times=seq(0,50,by=0.01), parms=pars)

ggplot(as.data.frame(res), aes(x,y, col=time))+geom_point()
ggplot(as.data.frame(res), aes(x,z, col=time))+geom_point()
ggplot(as.data.frame(res), aes(y,z, col=time))+geom_point()

ressimple = as.data.frame(res) %>% gather(key="var", value="value",-time)
ggplot(ressimple, aes(time, value, col=var))+geom_line()

# try with different initial conditions
pars = list(a=15,b=28,c=8/4)
res = ode(func=lorenz, c(x=0.3,y=5,z=10), times=seq(0,50,by=0.01), parms=pars)

ggplot(as.data.frame(res), aes(x,y, col=time))+geom_point()+scale_colour_gradientn(colours = terrain.colors(10))
ggplot(as.data.frame(res), aes(x,z, col=time))+geom_point()+scale_colour_gradientn(colours = terrain.colors(10))
ggplot(as.data.frame(res), aes(y,z, col=time))+geom_point()+scale_colour_gradientn(colours = terrain.colors(10))

ressimple = as.data.frame(res) %>% gather(key="var", value="value",-time)
ggplot(ressimple, aes(time, value, col=var))+geom_line()


```

