---
title: "predprey
"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(deSolve)
library(ggpubr)
library(tidyverse)
library(sensitivity)
library(lhs)
library(purrr)
library(ggpubr)
library(here)
```

## Example implementation

```{r}
source(here("R/lotvmod.R"))
lotvmod

# note the use of with
# initial conditions
currpop <- c(prey = 10, pred = 1)

# time points to see results
days <- seq(from = 1, to = 100, by = 1)

# set parameters
pars <- c(rprey = 0.5, alpha = 0.3, eff = 0.2, pmort = 0.2, K = 100)

# run the model
res <- ode(func = lotvmod, y = currpop, times = days, parms = pars)
```

## Run this model - how would you visualize results?

## Visualizing results

* two variable dynamic models that have feedbacks between variables can create cyclical
dynamics (and more complex )

* Two ways to look at results

  * time series of each state variable  (pred and prey)
  
  * how state variables interact with each other 
    * interactions through time
    * x versus y colored by time

## Relationship between populations {.scrollable}
```{r}
# graph the results
head(res)
# rearrange for easy plotting
resl <- as.data.frame(res) %>% pivot_longer(-time, names_to = "animal", values_to = "pop")
p1 <- ggplot(resl, aes(time, pop, col = animal)) +
  geom_line()

p1


p2 <- ggplot(as.data.frame(res), aes(pred, prey)) +
  geom_point() +
  labs(y = "Prey", x = "Predators")
p2

# To make this easier to understand - maybe
p2b <- ggplot(as.data.frame(res), aes(pred, prey, col = time)) +
  geom_point() +
  labs(y = "Prey", x = "Predators")
p2b

ggarrange(p1, p2b)
```

# Try other parameters {.scrollable}

- try to bring relative size of predators (versus prey) higher 

- what if you increase the predation rates (what might that look like in reality)

- how might you add a carrying capacity   


## Implementation {.scrollable}

```{r examples}
source(here("R/lotvmodK.R"))
lotvmodK

# initial conditions
currpop <- c(prey = 1, pred = 1)

# set parameter list
pars <- c(rprey = 0.1, alpha = 0.6, eff = 0.8, pmort = 0.4, K = 20)

# times when you want to evaluate
days <- seq(from = 1, to = 500)

# run our differential equation solver
res <- ode(func = lotvmodK, y = currpop, times = days, parms = pars)

# rearrange for plotting
resl <- as.data.frame(res) %>% pivot_longer(-time, names_to = "species", values_to = "pop")

# graph both populations over time
p1 <- ggplot(resl, aes(time, pop, col = species)) +
  geom_line()
p1

# also look at relationships between preditor and prey population and use color for time
# I will remove the legend here to make it easier to see
p2 <- ggplot(as.data.frame(res), aes(pred, prey, col = (round(time / 10)))) +
  geom_point() +
  theme(legend.position = "none")
p2
p2 <- ggplot(as.data.frame(res), aes(pred, prey, col = as.factor(round(time / 10)))) +
  geom_point() +
  theme(legend.position = "none")
p2
ggarrange(p1, p2)

# try with different parameter sets, can you create one where populations are stable - less cycling?
```


## Code for Lorenz system

Lets look at a Lorenz System Code
```{r lorenze}
# lorenz
source(here("R/lorenz.R"))

lorenz
```

## Application of Lorenz {.scrollable}


```{r lorenzeapp}
pars <- list(a = 10, b = 28, c = 8 / 3)
res <- ode(func = lorenz, c(x = 0.1, y = 0, z = 0), times = seq(0, 50, by = 0.01), parms = pars)
```
 now plot 


* as a phase space diagram (x vs y, x vs z, y vs z) - using time for color
* as a time series (e.g time on the x axis) add each variable with a different color

then try with different initial conditions

## Plots {.scrollable}

```{r, lr1}


ggplot(as.data.frame(res), aes(x, y, col = time)) +
  geom_point()
ggplot(as.data.frame(res), aes(x, z, col = time)) +
  geom_point()
ggplot(as.data.frame(res), aes(y, z, col = time)) +
  geom_point()

resl <- as.data.frame(res) %>% gather(key = "var", value = "value", -time)
ggplot(resl, aes(time, value, col = var)) +
  geom_line()
```

## Plot with different initial conditions
```{r, lr2}
# try with different initial conditions
pars <- list(a = 15, b = 28, c = 8 / 4)
res <- ode(func = lorenz, c(x = 0.3, y = 5, z = 10), times = seq(0, 50, by = 0.01), parms = pars)

ggplot(as.data.frame(res), aes(x, y, col = time)) +
  geom_point() +
  scale_colour_gradientn(colours = terrain.colors(10))
ggplot(as.data.frame(res), aes(x, z, col = time)) +
  geom_point() +
  scale_colour_gradientn(colours = terrain.colors(10))
ggplot(as.data.frame(res), aes(y, z, col = time)) +
  geom_point() +
  scale_colour_gradientn(colours = terrain.colors(10))

resl <- as.data.frame(res) %>% gather(key = "var", value = "value", -time)
ggplot(resl, aes(time, value, col = var)) +
  geom_line()
```


## Sensitivity analysis

Consider pred-prey 
BUT what will be the output - if we want to 'quantify sensitivity'
useful to look at a single value or set of value

For example

* Max Prey Population
* Min Prey Population


## Sensitivity Analysis steps

* Generate parameters (LHS, Sobol)
* Metrics function
* Wrapper Function
* Run wrapper function to get metrics for all parameter sets
* Graph and compute sensitivity statistics

 Discussion section will review how to do with LHS
 
## Example Sensitivity Analysis {.scrollable}

Given a Predator Prey model with the possible values for parameters

* $K$ - mean of 150, standard deviation 20
* $r_{prey}$ - some where between 0.01 and 0.3
* $\alpha$ - somewhere between 0.1 and 0.4
* $eff$ - mean 0.3 standard deviation 0.01
* $pmort$ - somewhere between 0.01 and 0.45

How do the maximum and minimum numbers of predators and prey vary across parameters; which is most important in controlling population dynamics?

Initial Conditions:  start with 1 predator and 1 prey and run for 500 time steps

```{r odesen}
source(here("R/lotvmodK.R"))
# lets start with sobol
library(sensitivity)


# want to learn about sensitivity to growth rate (r) and carrying capacity
# set the number of parameters
np <- 200
K <- rnorm(mean = 150, sd = 20, n = np)
rprey <- runif(min = 0.01, max = 0.3, n = np)
alpha <- runif(min = 0.1, max = 0.4, n = np)
eff <- rnorm(mean = 0.3, sd = 0.01, n = np)
pmort <- runif(min = 0.01, max = 0.45, n = np)

X1 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort)

# repeat to get our second set of samples
np <- 200
K <- rnorm(mean = 150, sd = 20, n = np)
rprey <- runif(min = 0.01, max = 0.3, n = np)
alpha <- runif(min = 0.1, max = 0.4, n = np)
eff <- rnorm(mean = 0.3, sd = 0.01, n = np)
pmort <- runif(min = 0.01, max = 0.45, n = np)

X2 <- cbind.data.frame(rprey = rprey, K = K, alpha = alpha, eff = eff, pmort = pmort)


# create our sobel object and get sets ofparameters for running the model
sens_PP <- sobolSalt(model = NULL, X1, X2, nboot = 300)

# name parameter sets...
colnames(sens_PP$X) <- c("rprey", "K", "alpha", "eff", "pmort")

# our metrics
# lets say we  want the maximum and minimum  of both predictor and prey

compute_metrics <- function(result) {
  maxprey <- max(result$prey)
  maxpred <- max(result$pred)
  minprey <- min(result$prey)
  minpred <- min(result$pred)
  return(list(maxprey = maxprey, minprey = minprey, maxpred = maxpred, minpred = minpred))
}

# build a wrapper function


p_wrapper <- function(rprey, alpha, eff, pmort, K, currpop, days, func) {
  parms <- list(rprey = rprey, alpha = alpha, eff = eff, pmort = pmort, K = K)
  result <- ode(y = currpop, times = days, func = func, parms = parms)
  colnames(result) <- c("time", "prey", "pred")
  # get metrics
  metrics <- compute_metrics(as.data.frame(result))
  return(metrics)
}

```

## Now run wrapper for all parameters {.scrollable}

* graph 
* sobol indices

```{r odesen2}
# run our model for all parameters and extract the results
currpop <- c(prey = 1, pred = 1)
days <- seq(from = 1, to = 500)
allresults <- as.data.frame(sens_PP$X) %>% pmap(p_wrapper, currpop = currpop, days = days, func = lotvmodK)

# take results back to unlisted form
allres <- allresults %>% map_dfr(`[`, c("maxprey", "minprey", "maxpred", "minpred"))


# range of response across parameter uncertainty
allresl <- allres %>% gather(key = "metric", value = "pop")
ggplot(allresl, aes(metric, pop)) +
  geom_boxplot()

# dealing with different scales
ggplot(allresl, aes(metric, pop, col = metric)) +
  geom_boxplot() +
  facet_wrap(~metric, scales = "free")
# plot cummulative densities

ggplot(allresl, aes(pop, col = metric)) +
  stat_ecdf(geom = "line") +
  facet_wrap(~metric, scales = "free")

# create sobol indices for Max Prey
sens_PP_maxprey <- sens_PP %>% sensitivity::tell(y = allres$maxprey)
rownames(sens_PP_maxprey$S) <- c("rprey", "K", "alpha", "eff", "pmort")
sens_PP_maxprey$S
rownames(sens_PP_maxprey$T) <- c("rprey", "K", "alpha", "eff", "pmort")
sens_PP_maxprey$T
```

## What does this tell you

* what are the most important parameters 
  *  maximum prey
  *  minimum prey
  *  maximum predator
  *  minimum predator
