---
title: "simpledynamic"
output: html_document
date: '2022-04-28'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(purrr)

```

# Implementing Dynamic Models in R

Dynammic models always involves derivatives (equations that express how things change from time step to time step or place to place )


Implement population **growth** as a derivative - a model of population change

```{r derivaties, echo=TRUE}

# note that we include time here but we don't use it; we will need this later
source("../R/dexppop.R")

# see how it works
dexppop(P=20, r=0.01)


#what is this?

# notices this is the same as
dexppop(t=100,P=20, r=0.01)

# lets look at this for a range of initial populations
pops = seq(from=1, to=100)
tmp = pops %>% map(~dexppop( time=0,r=0.01, P=.x))
pchange = unlist(tmp)


pdyn = data.frame(pops, pchange)
ggplot(pdyn, aes(pops, pchange))+geom_point(col="green", size=1.5)


# why is this a straight line?
# how many new individuals are born at each population level

# try this - add a carrying capacity (dP/dt = 0 if P > carryingcapacity)
```

# Integration

What if we wanted to look at population in 20 years given an initial condition

Two options

* explicit solution to differential equation is known; e.g. you can integrate both sides of the equation! Not always possible but lets look at a case where it is possible

* must be solved by iteration; this is what we do when we can't integrate both sides 

# Explicit Solution is available


```{r explicit}
source("../R/exppop.R")

# gives population after any time given an initial population

# 20 rabbits, growth rate of 0.01 how many in 30 years
exppop(T=30, P0=20, r=0.01, K=1000)

# if we want to see how population evolves over time - generate a time series by running our model for each point in time

initialrabbits = 20
years = seq(from=1, to=100, by=2)
Ptime = years %>% map_dbl(~exppop( P0=initialrabbits, r=0.01, K=1000, T=.x))

# keep track of what times we ran
Ptime = data.frame(P=Ptime, years=years)

ggplot(Ptime, aes(years,P))+geom_point()+labs(x="years",y="Rabbit Population")

# try generating results for maximum and minimum possible r values to compare (guess at what you think)


max_r = 0.1
min_r = 0.01
K = 1000

tmp = years  %>% map_dbl(~exppop(r=max_r, P0=initialrabbits, K=K, T=.x))
Ptime$Pmaxr = tmp
tmp = years %>% map_dbl(~exppop(r=min_r, P0=initialrabbits, K=K, T=.x))
Ptime$Pminr = tmp

head(Ptime)
Ptimep = Ptime %>% gather(key="r",value="P",-years)
ggplot(Ptimep, aes(years,P, col=r))+geom_point()+labs(x="years",y="Rabbit Population")

# notice how populatin becomes unstable for high growth rates! 


```
