---
title: "More Informal Sensitivity Analysis"
format: revealjs
editor: visual
---

## Sensitivity Analysis with *ethresh* {.scrollable}

There is a threshold radiation below which efficiency declines

-   In class we used *eff*

-   Assume +-15% uncertainty in *ethresh* parameter (uniform distribution)

-   Default value (from the function) is 10,000

-   Code so that its easy to change the uncertainty

-   Think about a data structure to allow us to see how uncertainty impacts annual power estimation for each year, or mean annual power generation

## First try on your own - then I'll go through the results

## Result

```{r ethresh, echo=TRUE}
# lets try sensitivity analysis again,
# ethresh - is defaulted to 10000; lets try +- 15%
# lets try 40 samples
# assume a uniform distribution - we don't know anything about variation
deviation <- 0.15
base_thresh <- 10000
ethresh <- runif(
  min = base_thresh - deviation * base_thresh,
  max = base_thresh + deviation * base_thresh, n = 40
)

# now run our model for the parameters
site2 <- ethresh %>% map(~ solarpv(
  area = 0.1,
  solar = sierraczosolar, clr = "green",
  eunit = "W", g = FALSE, etype = "direct", eff = 0.8, PR = 0.75,
  ethresh = .x
))


# extract a useful data structure,lets say we want  just the annual data (not the mean), and then reformat as a data frame with nice column names
tmp <- map_df(site2, `[`, c("annual"))
site2df <- data.frame(
  year = tmp$annual$year,
  elect = tmp$annual$elect
)

# now we could plot
ggplot(
  site2df,
  aes(as.factor(year), elect, group = year)
) +
  geom_boxplot() +
  labs(y = "Electricity generated in W", x = "Year")

# we also might want to see mean electricity generation across parameter uncertainty
mean_elect <- map_df(site2, `[`, c("mean"))

# because there is only 1 value for each ethresh we can add as a column
mean_elect$ethresh <- ethresh


p1 <- ggplot(mean_elect, aes(y=mean)) +
  geom_boxplot() +
  labs(x = "Mean Annual Electricity W")
p1

# or we might want to see how electricity varies with parameter value
p2 <- ggplot(mean_elect, aes(ethresh, mean)) +
  geom_point(cex = 2, col = "orange") +
  labs(y = "Mean Annual Electricity W", x = "Threshold Radiation (kJ/m2)  \nabove which energy production is more efficient")
p2
```

## Sensitivity to Energy Threshold is not linear

Sensitivity to parameters that effect a theshold is often complex (and not linear!)

```{r, echo=TRUE}
ggarrange(p1, p2)
```

## Something to think about- what matters more uncertainty in *ethresh* or *eff*?

learn more about Quarto presentations see <https://quarto.org/docs/presentations/>.
