---
title: "Discussion5_Summarizing Dynamic Model Output"
format: revealjs
theme: solarized
resources: ["img/"]
css: ["slides.css"]
editor: visual
---

## Extracting information from space-time results {.scrollable}

In class we've talking about summarizing results from dynamic models

-   pictures can be hard to interpret

-   summarizing over one of the dimensions (either space or time) can help

-   looking at a single trajectory through time

-   looking at spatial variation for one point in time

-   looking at spatial variation for multiple points in time

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(deSolve)
library(here)
```

## Extracting from our diffusion model {.scrollable}

```{r}
source(here("R/diffusion.R"))
resfast_fixtime <- diff1(initialC = 100, nx = 10, dx = 0.5, nt = 100, dt = 0.5, D = 0.2, area = 1)


filled.contour(resfast_fixtime$conc, xlab = "time", ylab = "Distance Along Path", main = "Concentration through time and space")
```

# Summarizing Model Output

-   Summarize relative to what you care about

Lets say this was a pollutant injected into a stream...

-   lets say you have a vulnerable habit located 3 spatial units from the injection point

-   lets see what its pollutant exposure looks like.\

```{r one option}
# View(resfast_fixtime$conc)


# graph a single point in space through time
# single column (time)
plot(resfast_fixtime$conc[, 3], ylab = "Concentration for a location 2 spatial units from origin (1+2")
```

-   what if we were concerned about all locations

```{r all locations}
# plot all trajectories
# add a time column to concentration data frame and transform for plotting
resl <- as.data.frame(resfast_fixtime$conc) %>%
  mutate(time = seq(from = 1, to = 100)) %>%
  pivot_longer(-time, names_to = "distance", values_to = "conc")
ggplot(resl, aes(time, conc, col = distance)) +
  geom_line()

# plot all places at each point in time
ggplot(resl, aes(time, conc, group = time)) +
  geom_boxplot()
```

# Spatial summaries

-   interested in how quickly the pollutant evens out through space

```{r spatial summaries}

# use apply to calculate the spatial variation for each row (e.g for each time point)
cvar <- resfast_fixtime$conc %>% apply(1, var)
cmean <- resfast_fixtime$conc %>% apply(1, mean)

spatial_aver <- cbind.data.frame(cvar, cmean, time = seq(from = 1, to = 100))
length(cvar)
# notice its the same as the number of time units (nt) used above

# plot spatial variation through time
ggplot(spatial_aver, aes(time, cvar)) +
  geom_line() +
  labs(y = "Spatial Variation")

# plot coefficient of variation (so standard deviation divided by the mean)
ggplot(spatial_aver, aes(time, 100 * sqrt(cvar) / cmean)) +
  geom_line() +
  labs(y = "COV (as percent")
```

## it is helpful to turn the summary into a function that you can apply to the dynamic models output

## example

## In class exerise {.scrollable}

For the diffusion model that we've been working with

-   Try to create a function that counts how many times (any where in the stream concentrations are above some threshold)
