---
title: "Continuing with dynamic models"
format: revealjs
theme: solarized
resources: ["img/"]
css: ["slides.css"]
editor: visual
---



# Application of diffusion model

```{r, rundiffusion}
source(here("R/diffusion.R"))

# run our diffusion model (iterative difference equation) with initial concentration of 10, for 8 timestep (size 1m), and 10 space steps (size 1s)
# using diffusion parameters 0.5 s/m2, 10 m2

# take a look at the list  returned with our 3 data frames for concentration (conc), qin and qout


# used filled contour to plot results

# or if you prefer this orientation (Distance on x axis)

# you can also plot how much material is moving in to each cell at each time step

```

## Change parameters (diffusivity D, and space and time steps (dx, dt)) {.scrollable}

Play on your own or with a partner

Try running the diffusion model with different time steps, space steps and parameters

## Example {.scrollable}

```{r, alternativediffusion}
# what if we increase diffusivity
resfast <- diff1(initialC = 100, nx = 10, dx = 0.5, nt = 10, dt = 10, D = 0.08, area = 1)
filled.contour(resfast$conc, xlab = "Time", ylab = "Distance", main = "Concentration through time and space")
filled.contour(resfast$qin, xlab = "Time", ylab = "Distance", main = "Qin")
```

## Fixing the problem {.scrollable}

-   difference equations as an approximation (and the challenges that methods for numerical integration try to overcome)
-   if things are changing quickly we need to use:
    -   much smaller time,
    -   much smaller space steps to avoid overshoot and instability

## Solution {.scrollable}

```{r, fixed}
# so lets cut our step size by 10 (dt=1) (but then  multiply number of steps (nx=10) to cover the same distance)
# initialC is still 100, diffusivity D is still 0.08, area is still 1

# run the model with these parameters and plot results

# use filled contour to plot concentration
filled.contour(resfast_fixtime$conc, xlab = "time", ylab = "Distance Along Path", main = "Concentration through time and space")

# use filled contour to plot qin and qout

# how would you plot net transport?

```

## Dynamics Models {.scrollable}

-   Diffusion example illustrates the challenge of numerical integration

-   We see evidence of "overshoot"

-   Correct by reducing the time step (but then we have to increase the number of time steps to cover the same period)

-   This is a common issue with numerical integration of dynamic models (models with feedbacks)

-   Many methods to try to overcome this issue \* Euler \* Runge-Kutta, \* Adams-Bashforth-Moulton

-   When you run solvers (like *ode*) they use these methods

## Extracting information from space-time results {.scrollable}

-   pictures/maps can be hard to interpret

-   summarizing over one of the dimensions (either space or time) can help

    -   looking at a single trajectory through time

    -   looking at spatial variation for one point in time

    -   looking at spatial variation for multiple points in time

## Examples from diffusion {.scrollable}

```{r option}
# View(resfast_fixtime$conc)


# graph a single point in space through time
# single column (time)

# plot all trajectories
# add a time column to concentration data frame and transform for plotting

# use apply to calculate the spatial variation for each row (e.g for each time point)

# combine into a data frame for plotting

# notice its the same as the number of time units (nt) used above

# plot spatial variation through time

# plot coefficient of variation (so standard deviation divided by the mean)

```

## Extracting meaning from time series output {.scrollable}

Ok these graphs help but what if we need a single metric to summarize results from diffusion model - would depend on the application

Quiz - Suggest a metric (a single measure)

## Summary Measures

-   time it takes for concentration to reach a certain threshold at a particular location
-   time it takes for all cells to have a value less than a threshold
-   maximum concentration at a particular location
-   time it takes for the mean concentration across all cells be less than some threshold

## One example

time for all cells to be less than a threshold

lets say a dangerous threshold is 21

```{r summarydynamic}
# recall we have a data frame with concentration for each cell at each time point


# find the first time point where all cells are less than 21


# as percent of simulation time

```

## Dynamic models {.scrollable}

-   models that have feedbacks (conditions evolve through time)

-   numerical integration - usually done with a solver

    -   only one independent variable ordinary differential equation (e.g just time)
    -   we can use the ODE solver
    -   derivative is first order
    -   (e.g $dy/dt$ = ; not $d^2y/dt$+dy/dy = f(y,t))
    -   there also solvers for high order and partial differential equations

## Sensitivity Analysis of a Differential Equation {.scrollable}

We can apply sensitivity analysis to a differential equation

A key issue where is sensitivity of what?

Dynamic models often give you many many outputs

-   time series (streamflow every day for a year, population for 30 years)
-   or output over space (spatially averaged concentration after 10 days?)

So if we are asking 'sensitivity of what' we need to summarize results in some way (reduce their dimensionality )

## Some options for reducing output dimensionality (summarizing output) {.scrollable}

Depends on what is important for your model application

-   max
-   mean
-   min
-   total
-   variation
-   time it takes for something to happen

So a key step in sensitivity analysis with a dynamics model is summarizing results into a few key measures

Its useful to turn that summarizing workflow into a function

## Practice

write a function that extracts a summary metric from diffusion model output

## Workflow for Sensitivity Analysis {.scrollable}

1)  implement (or identify pre-existing) dynamic model

2)  obtain samples of parameter sets (from sobel or LHS)

3)  build a function that will extract the information (metrics) you want from your dynamic model (output of the ode)

4)  create a data structure to store the metrics for each parameter set - in my example I call it metrics (but could be anything)

5)  decide on initial conditions and time period over which you will run the model

6)  run ODE for each parameter sets to fill in this metrics data structure

    -   its usually helpful to create a wrapper function
        -   runs ODE
        -   extracts metrics
    -   run wrapper function for each parameter sets

7)  send the metrics back to the sensitivity analysis object (from sobel or LHS)

8)  plot and analyze results
