---
title: "Lecture 17 Calibration2"
format: revealjs
execute: 
  echo: TRUE
theme: solarized
resources: ["img/"]
css: ["slides.css"]
editor: visual
---

## Glue - generalized uncertainty analysis {.scrollable}

What if we wanted to keep all of the 'good' parameters

-   we could just keep them all as equally likely
-   we could weight them by performance

Either way we can graph and come up with 'best' prediction accounting for uncertainty

Create a single measure of accuracy 


* *compute_lowflowmetrics_all* added a summary measures 

-   relative error in annual minimum flow estimate
-   relative error in monthly flow during low flow period
-   correlation between observed and modelled annual minimum flow
-   correlation between observed and modelled flow during the low flow period


Updates (from *compute_lowflowmetrics*) 

* allows user to input weights (default evenly weighted)
* transforms error based metrics into 0-1 scale
```{r startup}

library(tidyverse)
library(here)

# read metric function
source(here("R/compute_lowflowmetrics_all.R"))

# load model results from last class
msage = readRDS(here("Data/msage.RDS"))

# first rearrange so we can plot all results
msagel = msage %>% gather(key="sim",value="str", -date, -month, -day, -year, -wy,-obs)

# apply our accurracy measure to each parameter set 
res <-  msage %>% select(-date, -month, -day, -year, -wy, -obs ) %>%
  apply(2, compute_lowflowmetrics_all, o = msage$obs,
              month=msage$month, year=msage$year, day=msage$day, wy=msage$wy) 
results = as.data.frame(matrix(unlist(res), byrow=T, ncol=5))
colnames(results)=c("annual_min_err","annual_min_cor", "low_month_err", "low_month_cor", "combined")
results$sim <- msage %>% select(-date, -month, -day, -year, -wy, -obs ) %>% colnames()

```


## Use the accuracy measure {.scrollable}

Use the combined accuacy measure to define behavioural (acceptable) parameter set (**results_acc**) - two options

* define a threshold (we will use 30%)
* take top 50 performing parameter sets


```{r behavioral, echo=TRUE, eval=TRUE}
summary(results$combined)

# 1) selecting behaviorial or acceptable parameters sets

threshold <- 0.3
results_acc <- subset(results, combined > threshold)
head(results_acc)

# as an alternative  what if you want the top N parameter sets
topN <- 50
tmp <- results[order(results$combined, decreasing = T), ]
results_acc <- tmp[1:topN, ]
head(results_acc)
```

## Defining weights (likelihood) for parameter sets {.scrollable}

Now define "weights" (likelihood) based on parameter performance for the acceptable or behaviorial parameters

* accuracy measure defined above will define weight
* sum of the weights to equal 1
* we normalize by the range of accuracy for the behavioral parameters
  * this **relative accuracy** becomes the weight
* we now only work with behavioural parameter sets 
 
To normalize

$weight$ = $\frac{accuracy - min(accuracy)}{max(accuracy) - min(accuracy)}$

*weight* will be from 0-1 

To make sure it sums to 1 we can divide by the sum of all weights

$weight$ = $\frac{weight}{\sum weight}$


## Weighting in R {.scrollable}

```{r weighting, echo=TRUE, eval=TRUE}
# create a weight for each parameter set based on its relative accuracy - we do this so all weights sum to 1
max_acc <- max(results_acc$combined)
min_acc <- min(results_acc$combined)

# normalize weights
results_acc$w_acc <- (results_acc$combined - min_acc) / (max_acc - min_acc)

# make sure weights sum to one
sum_acc <- sum(results_acc$combined)
results_acc$wt_acc <- results_acc$combined / sum_acc

# look at values
summary(results_acc$wt_acc)
# check to see that they sum to one
sum(results_acc$wt_acc)

Nacc <- nrow(results_acc)
Nacc
```

# Using weights {.scrollable}

Use weights is to define a maximum likelihood estimate
* by averaging (weighted by accuracy) streamflow from all behavioural simulations

$MLE$ = $\frac{\sum_{i=1}^{N} flow_i \times weight_i}{\sum_{i=1}^{N} weight_i}$

where 
* $flow_i$ is the streamflow from simulation $i$ 
* $weight_i$ is the weight for that simulation

* because we know weights sum to 1 we can just multiply the flow by the weight and sum

## MLE in R {.scrollable}
```{r mle, echo=TRUE, eval=TRUE}
# generate streamflow as weighted average of all  acceptable parameter sets


# subset only acceptable runs
msagel_acc <- subset(msagel, sim %in% results_acc$sim)
# join with weights from res_acc, left_join will repeat weights for each day in streamflow trajectory
msagel_acc <- left_join(msagel_acc, results_acc, by = "sim")
head(msagel_acc)
# finally multiply flow by weight
msagel_acc <- msagel_acc %>% mutate(str_wt = str * wt_acc)

# now we can average streamflow for each day from all the runs # using the weights
aver_flow <- msagel_acc %>%
  group_by(date) %>%
  dplyr::summarize(str_mse = sum(str_wt))

# add some date information 
aver_flow <- left_join(aver_flow, msage[, c("date", "month", "year", "day", "wy", "obs")],
  by = c("date")
)

# plot
aver_flowl = aver_flow %>% pivot_longer(
  cols = c(str_mse, obs),
  values_to = "str", names_to = "type"
)
```

# Graphs of MLE {.scrollable}

```{r mle2, echo=TRUE, eval=TRUE}
ggplot(aver_flowl, aes(x = date, y = str, col=type)) +
  geom_line()+
  labs(y = "Streamflow mm/day")

# add some of the other date info and plot a subset
wycheck <- 1985
ggplot(subset(aver_flowl, wy == wycheck), aes(x = date, y = str, col=type)) +  labs(y = "Streamflow mm/day") + geom_line()
```

# Monthly flow with MLE {.scrollable}

```{r mlemonth, echo=TRUE}
# we could also calculate mse for august flow
# get monthly flow

msagel_acc_month <- msagel_acc %>%
  group_by(month, wy, sim) %>%
  dplyr::summarize(str = sum(str), obs = sum(obs), wt_acc = mean(wt_acc))

# now MSE for monthly flow
mse_flow <- msagel_acc_month %>%  
  group_by(wy, month) %>%
  dplyr::summarize(str_mse = sum(str*wt_acc), obs = mean(obs))

```

# Graph Monthly MLE {.scrollable}

```{r mlemonth2, echo=TRUE}
p1=ggplot(subset(msagel_acc_month, month==8), aes(wy, str, group=wy)) +
  geom_boxplot(position="dodge") +
  geom_point(aes(wy, obs), col="red") +
  labs(y = "Streamflow", x = "Date")

p1 + geom_point(data=subset(mse_flow, month==8), aes(wy, str_mse), col="blue") 


```

