---
title: "Sensitivity Analysis with Sobol"
format: revealjs
theme: solarized
resources: ["img/"]
css: ["slides.css"]
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(sensitivity)
library(tidyverse)
# library(lhs)
library(purrr)
library(here)
```

## Larger Complex Models {.scrollable}

Only do sensitivity on some inputs/parameters

Use Latin Hypercube to generate samples

Run model for samples

Generate Summary Statistics and graph

## More Information on Sensitivity Analysis {.scrollable}

[Applying Sensitivity Analysis in Biology](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2570191/)

(focuses on both LH and eFAST (another variance based sensitivity analysis like Sobol))

Nice papers on sensitivity analysis for environmental modeling

-   Saltelli, Andrea, and Paola Annoni. "How to avoid a perfunctory sensitivity analysis." Environmental Modelling & Software 25, no. 12 (2010): 1508-1517

-   Tarantola, Stefano, Federico Ferretti, Samuele Lo Piano, Mariia Kozlova, Alessio Lachi, Rossana Rosati, Arnald Puy et al. "An annotated timeline of sensitivity analysis." Environmental Modelling & Software 174 (2024): 105977.

------------------------------------------------------------------------

## Steps in Sensitivity Analysis {.scrollable}

Define the distribution of input parameters (e.g normal with its mean and standard deviation)

Define the outputs to be considered (e.g if your output is time series are you looking at daily, max, min, annual ?)

Sample from pdf of input parameters - we use *randomLHS* for this

Run the model for each sample

Graph the results

Quantify sensitivity - we used *pcc*

------------------------------------------------------------------------

## Sobol {.scrollable}

Sobol sensitivity analysis is similar to LHS approach as a way to efficiently sample parameter space

Sobol is more general - it can use measured parameter samples (we will approximate sampling here by using known distributions but you could use actual samples!)

Sobol is a *variance-based* method

Sobol quantifies sensitivity by breaking the variance of output into contributions by variation in parameters

## Sobol Indices - Quantifying Sensitivity {.scrollable}

Indices are computed using parameter-output variance relationships; they are estimates so the indices themselves have uncertainty bounds

Several Indices

[First order sensitivity or Main Effect]{style="color: green;"}

-   variance associated directly with parameter alone
-   fraction associated with each parameter, sum to 1 (although because it is estimated can be slightly more or less)

[Total Effect]{style="color: green;"}

-   variance associated with parameter and interaction with other parameters
-   sum can be more than 1 if parameters interact

[Second Order Indices]{style="color: green;"}

-   less used but quantify how parameter pair-wise parameter interactions contribute to output variation

## Sobol in R {.scrollable}

-   *Sensitivity* package

-   Sobol Indices require estimation - and there are different methods to do that

-   The *Sensitivity* package has several of those

-   today we will use *sobolSalt* - which uses a method by Saltelli (who has written extensively on Sensitivity analysis)

-   R help pages for *Sensitivity* provide many good references

## Sobol - how to {.scrollable}

-   run Sobol to get parameter sets in a sensitivity analysis object
-   run model with those parameter sets
-   *tell* the senstivity object about results associated with each parameter set
-   look at sensitivity analysis indices from Sobol

Generation of parameter sets slightly different

-   generate **two** samples of parameter sets by sampling from a-priori (expected) distributions
-   these would be the "hypothetical" distributions based on assumptions about the data
-   ideally these would be distributions that you actually sampled

## Example Atmospheric Conductance {.scrollable}

Often when we are estimating vegetation or crop water use we need to know the atmospheric conductance - which is essentially how easily water diffuses into the air and depends largely on windspeed (you get more evaporation in windier conditions)

Atmospheric conductance is also influenced by the vegetation itself and the turbulence it creates

I've provided a function to compute atmospheric conductance $C_{at}$ (how easily vapor diffuses from vegetation surfaces)

**The function *Catm.R* is provided**

## Atmospheric conductance (simple model) {.scrollable}

So that you know what it does - here's some background on the function $$
C_{at} = \frac{v_m}{6.25*{ln(\frac{z_m-z_d}{z_0})}^2}
$$ $$
z_d = k_d*h
$$ $$
z_0 = k_0*h
$$

$z_m$ is the height at which windspeed is measured - must be higher than the vegetation (cm), it is usually measured 200 cm above the vegetation

$h$ is vegetation height (cm)

$v$ is windspeed (cm/s)

Typical values if $k_d$ and $k_o$ are 0.7 and 0.1 respectively (so use those as defaults)

## Sensitivity Anallysis Goal {.scrollable}

For a given forest, perform a sensitivity analysis of model predictions of conductance. Consider the sensitivity to uncertainty in the following parameters and inputs

-   $height$

-   $k_d$

-   $k_0$

-   $v$

Windspeeds $v$ are normally distributed with a mean of 250 cm/s with a standard deviation of 30 cm/s

For vegetation height assume that height is somewhere between 9.5 and 10.5 m (but any value in that range is equally likely)

For the $k_d$ and $k_0$ parameters you can assume that they are normally distributed with standard deviation of 1% of their default values

## Quiz {.scrollable}

Use intuition to guess which parameter has the highest **first-order sensitivity**?

Varying: $h$, $v$, $k_d$, $k_0$

$$
C_{at} = \frac{v_m}{6.25\,[\ln((z_m-z_d)/z_0)]^2},
\quad
z_d = k_d h,\;
z_0 = k_0 h
$$

-   $k_0, k_d$: Normal, SD = 10% of mean\
-   $v$: Normal ($\mu=250$, $\sigma=30$)\
-   $h$: Uniform (9.5–10.5)

## Steps: Generate Parameters {.scrollable}

First we need to generate parameter sets for the Sobol analysis

-   2 sets of samples of all of the parameters, we will make these up by sampling from expected distributions
-   use *sobolSalt* to generate the parameter sets

```{r sobolpar}
source(here("R/Catm.R"))

# generate two examples of random number from parameter distributions

np <- 1000
k_o <- rnorm(mean = 0.1, sd = 0.1 * 0.1, n = np)
k_d <- rnorm(mean = 0.7, sd = 0.7 * 0.1, n = np)
v <- rnorm(mean = 250, sd = 30, n = np)
height <- runif(min = 9.5, max = 10.5, n = np)

X1 <- cbind.data.frame(k_o, k_d, v, height = height)

# repeat sampling
k_o <- rnorm(mean = 0.1, sd = 0.1 * 0.1, n = np)
k_d <- rnorm(mean = 0.7, sd = 0.7 * 0.1, n = np)
v <- rnorm(mean = 250, sd = 30, n = np)
height <- runif(min = 9.5, max = 10.5, n = np)

X2 <- cbind.data.frame(k_o, k_d, v, height = height)

# there are different versions of sobol functions that have different approaches for estimating parameters and indices, we use an approach implemented by jansen

sens_Catm_Sobol <- sobolSalt(model = NULL, X1, X2, nboot = 100)


# Take a look at the Sobol generated sensitivity object
# your parameters sets for sensitivity analysis are in X
```

## Steps: Compute Sobol Indices {.scrollable}

now run model for Sobol generated parameter sets and compute indices

```{r sobolrun}
# run model for all parameter sets
# make sure you give the parameters names

parms <- as.data.frame(sens_Catm_Sobol$X)
colnames(parms) <- colnames(X1)
res <- pmap_dbl(parms, Catm)


sens_Catm_Sobol <- sensitivity::tell(sens_Catm_Sobol, res, res.names = "ga")

# main effect:  partitions variance (main effect without co-variance) - sums approximately to one
sens_Catm_Sobol$S
# useful to add names
row.names(sens_Catm_Sobol$S) <- colnames(parms)
sens_Catm_Sobol$S

# total effect - accounts for parameter interactions
row.names(sens_Catm_Sobol$T) <- colnames(parms)
sens_Catm_Sobol$T

# Both the main effect and total effect can tell us something about how the parameter influences results


print(sens_Catm_Sobol)
```

## Steps: Interpreting Sobol Indices {.scrollable}

-   pay attention to values of the indices and confidence intervals
    -   if 0 is within the confidence interval, parameter uncertainty is not influencing output
-   substantial differences between total effect and first order indices suggest parameter interactions

TIP: a useful plotting strategy is to plot model output against parameter with the highest total effect and then use the parameter with second highest total effect for color

## Steps: Plotting {.scrollable}

-   uncertainty in the output

-   parameter-output variation for the most sensitive parameters

```{r sobolplot}
# graph two most sensitive parameters
both <- cbind.data.frame(parms, gs = sens_Catm_Sobol$y)

# look at overall gs sensitvity to uncertainty
ggplot(both, aes(x = gs)) +
  geom_histogram() +
  geom_vline(xintercept = mean(both$gs), col = "cyan")

# look at response of conductance to the two interesting variables
ggplot(both, aes(v, gs, col = height)) +
  geom_point() +
  labs(y = "Conductance (mm/s)", x = "Windspeed")


# look at response of conductance to the two most important variables
ggplot(both, aes(k_d, gs, col = k_o)) +
  geom_point() +
  labs(y = "Conductance (mm/s)", x = "k_d parameter")
# use second most sensitive parameter (using most important as color)
ggplot(both, aes(k_o, gs, col = k_d)) +
  geom_point() +
  labs(y = "Conductance (mm/s)", x = "k_d parameter")
```

## Second order indices {.scrollable}

Optional for this course

If you want to also compute a second order indices you need to use a different variation. (scheme=B)

There are multiple implementation of Sobol in the *sensitivity* package, they can be more or less stable I find *sobolSalt* works well

```{r, alternative}
sens_Catm_Sobol2 <- sobolSalt(model = NULL, X1, X2, nboot = 100, scheme = "B")

parms <- as.data.frame(sens_Catm_Sobol2$X)
colnames(parms) <- colnames(X1)
res <- pmap_dbl(parms, Catm)


sens_Catm_Sobol2 <- sensitivity::tell(sens_Catm_Sobol2, res, res.names = "ga")

# main effect:  partitions variance (main effect without co-variance) - sums approximately to one
row.names(sens_Catm_Sobol2$S) <- colnames(parms)
sens_Catm_Sobol2$S

# total effect - accounts for parameter interactions
row.names(sens_Catm_Sobol2$T) <- colnames(parms)
sens_Catm_Sobol2$T

# second order parameters interaction in controlling sensitivity
# parameters are in order, interactiosn are small here
sens_Catm_Sobol2$S2
```

# Assignment 5

Sobol Sensitivity Analysis ["../assignments/Assignment5_sobol.qmd"](../assignments/Assignment5_sobol.qmd)
