---
title: "Lecture 17 Calibration2"
format: revealjs
execute: 
  echo: TRUE
theme: solarized
resources: ["img/"]
css: ["slides.css"]
editor: visual
---

## Glue - generalized uncertainty analysis {.scrollable}

What if we wanted to keep all of the 'good' parameters

-   we could just keep them all as equally likely
-   we could weight them by performance

Either way we can graph and come up with 'best' prediction accounting for uncertainty

Create a single measure of accuracy

-   *compute_lowflowmetrics_all* added a summary measures

-   relative error in annual minimum flow estimate

-   relative error in monthly flow during low flow period

-   correlation between observed and modelled annual minimum flow

-   correlation between observed and modelled flow during the low flow period

Updates (from *compute_lowflowmetrics*)

-   allows user to input weights (default evenly weighted)
-   transforms error based metrics into 0-1 scale

```{r startup}

library(tidyverse)
library(here)

# read metric function
source(here("R/compute_lowflowmetrics_all.R"))

# load model results from last class
msage = readRDS(here("Data/msage.RDS"))

# first rearrange so we can plot all results
msagel = msage %>% gather(key="sim",value="str", -date, -month, -day, -year, -wy,-obs)

# apply our accurracy measure to each parameter set 
res <-  msage %>% select(-date, -month, -day, -year, -wy, -obs ) %>%
  apply(2, compute_lowflowmetrics_all, o = msage$obs,
              month=msage$month, year=msage$year, day=msage$day, wy=msage$wy) 
results = as.data.frame(matrix(unlist(res), byrow=T, ncol=5))
colnames(results)=c("annual_min_err","annual_min_cor", "low_month_err", "low_month_cor", "combined")
results$sim <- msage %>% select(-date, -month, -day, -year, -wy, -obs ) %>% colnames()

```

## Use the accuracy measure {.scrollable}

Use the combined accuacy measure to define behavioural (acceptable) parameter set (**results_acc**) - two options

-   define a threshold (we will use 30%)
-   take top 50 performing parameter sets

```{r behavioral, echo=TRUE, eval=TRUE}
summary(results$combined)

# 1) selecting behaviorial or acceptable parameters sets

threshold <- 0.3
results_acc <- subset(results, combined > threshold)
head(results_acc)

# as an alternative  what if you want the top N parameter sets
topN <- 50
tmp <- results[order(results$combined, decreasing = T), ]
results_acc <- tmp[1:topN, ]
head(results_acc)
```

## Defining weights (likelihood) for parameter sets {.scrollable}

Now define "weights" (likelihood) based on parameter performance for the acceptable or behaviorial parameters

-   accuracy measure defined above will define weight
-   sum of the weights to equal 1
-   we normalize by the range of accuracy for the behavioral parameters
    -   this **relative accuracy** becomes the weight
-   we now only work with behavioural parameter sets

To normalize

$weight$ = $\frac{accuracy - min(accuracy)}{max(accuracy) - min(accuracy)}$

*weight* will be from 0-1

To make sure it sums to 1 we can divide by the sum of all weights

$weight$ = $\frac{weight}{\sum weight}$

## Weighting in R {.scrollable}

```{r weighting, echo=TRUE, eval=TRUE}
# create a weight for each parameter set based on its relative accuracy - we do this so all weights sum to 1
max_acc <- max(results_acc$combined)
min_acc <- min(results_acc$combined)

# normalize weights
results_acc$w_acc <- (results_acc$combined - min_acc) / (max_acc - min_acc)

# make sure weights sum to one
sum_acc <- sum(results_acc$combined)
results_acc$wt_acc <- results_acc$combined / sum_acc

# look at values
summary(results_acc$wt_acc)
# check to see that they sum to one
sum(results_acc$wt_acc)

Nacc <- nrow(results_acc)
Nacc
```

# Using weights {.scrollable}

Use weights is to define a maximum likelihood estimate \* by averaging (weighted by accuracy) streamflow from all behavioural simulations

$MLE$ = $\frac{\sum_{i=1}^{N} flow_i \times weight_i}{\sum_{i=1}^{N} weight_i}$

where \* $flow_i$ is the streamflow from simulation $i$ \* $weight_i$ is the weight for that simulation

-   because we know weights sum to 1 we can just multiply the flow by the weight and sum

## MLE in R {.scrollable}

```{r mle, echo=TRUE, eval=TRUE}
# generate streamflow as weighted average of all  acceptable parameter sets


# subset only acceptable runs
msagel_acc <- subset(msagel, sim %in% results_acc$sim)
# join with weights from res_acc, left_join will repeat weights for each day in streamflow trajectory
msagel_acc <- left_join(msagel_acc, results_acc, by = "sim")
head(msagel_acc)
# finally multiply flow by weight
msagel_acc <- msagel_acc %>% mutate(str_wt = str * wt_acc)

# now we can average streamflow for each day from all the runs # using the weights
aver_flow <- msagel_acc %>%
  group_by(date) %>%
  dplyr::summarize(str_mse = sum(str_wt))

# add some date information 
aver_flow <- left_join(aver_flow, msage[, c("date", "month", "year", "day", "wy", "obs")],
  by = c("date")
)

# plot
aver_flowl = aver_flow %>% pivot_longer(
  cols = c(str_mse, obs),
  values_to = "str", names_to = "type"
)
```

# Graphs of MLE {.scrollable}

```{r mle2, echo=TRUE, eval=TRUE}
ggplot(aver_flowl, aes(x = date, y = str, col=type)) +
  geom_line()+
  labs(y = "Streamflow mm/day")

# add some of the other date info and plot a subset
wycheck <- 1985
ggplot(subset(aver_flowl, wy == wycheck), aes(x = date, y = str, col=type)) +  labs(y = "Streamflow mm/day") + geom_line()
```

# Monthly flow with MLE {.scrollable}

```{r mlemonth, echo=TRUE}
# we could also calculate mse for august flow
# get monthly flow

# summarize monthly flow
msagel_month <- msagel %>%
  group_by(month, wy, sim) %>%
  dplyr::summarize(str = sum(str), obs = sum(obs))

# extract only acceptable runs

msagel_acc_month <- subset(msagel_month, sim %in% results_acc$sim)
# join with weights from res_acc, left_join will repeat weights for each month in streamflow trajectory
msagel_acc_month <- left_join(msagel_acc_month, results_acc, by = "sim")


# now MSE for monthly flow
mse_flow <- msagel_acc_month %>%  
  group_by(wy, month) %>%
  dplyr::summarize(str_mse = sum(str*wt_acc), obs = mean(obs))

# For comparison, compute MSE for all simulations
mse_flow_all <- msagel_month %>%
  group_by(wy, month) %>%
  dplyr::summarize(str_mse = mean(str))

mse_flow_all = left_join(mse_flow_all, mse_flow, by = c("wy", "month"), suffix = c("_all", "_mse"))

# see the impact of calibration
tmp = mse_flow_all %>% subset(month==8) 
cor(tmp$str_mse_all, tmp$obs)
cor(tmp$str_mse_mse, tmp$obs)

# but what about across all months
cor(mse_flow_all$str_mse_all, mse_flow_all$obs)
cor(mse_flow_all$str_mse_mse, mse_flow_all$obs)
```

# Graph Monthly MLE {.scrollable}

```{r mlemonth2, echo=TRUE}
p1=ggplot(subset(msagel_acc_month, month==8), aes(wy, str, group=wy)) +
  geom_boxplot(position="dodge") +
  geom_point(aes(wy, obs), col="red") +
  labs(y = "Streamflow", x = "Date")

p1 + geom_point(data=subset(mse_flow, month==8), aes(wy, str_mse), col="blue") 


```

# Compare with results from last class {.scrollable}

```{r compare, echo=TRUE}

# plot mean August streamflow for all parameters and acceptable parameters
msagel_month_avg <- msagel_month %>%
  group_by(month, sim) %>%
  dplyr::summarize(str = mean(str), obs = mean(obs))
msagel_month_avg = msagel_month_avg %>% mutate(accept = ifelse(sim %in% results_acc$sim, "Accept", "Not"))

# read in the results from last class
goodpar = c(40, 42,29, 68,9,20, 12,32, 10,77)
other_cal_parms = paste0("V",goodpar)
msagel_month_avg$othercal = ifelse(msagel_month_avg$sim %in% other_cal_parms, "Accept B", "Not B")

# compute average august flow in observed
mean_august_obs = mean(subset(msagel_month_avg, month==8)$obs)
```

## Plot

```{r compare2, echo=TRUE}
p2 = ggplot(subset(msagel_month_avg, month==8), aes(x=accept, y=str))+geom_boxplot()+geom_hline(yintercept=mean_august_obs, col="red")+geom_boxplot(aes(x=othercal, y=str))  + labs(y = "Mean Streamflow\n", x = "Parameter set") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2
```

## Try a different month {.scrollable}

```{r different_month, echo=TRUE}

# compute average december flow in observed
mean_dec_obs = mean(subset(msagel_month_avg, month==12)$obs)

p3 = ggplot(subset(msagel_month_avg, month==12), aes(x=accept, y=str))+geom_boxplot()+geom_hline(yintercept=mean_dec_obs, col="red")+geom_boxplot(aes(x=othercal, y=str))  + labs(y = "Mean Streamflow \nDecember", x = "Parameter set") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3
```

## Annual Total or Minimumm Flow {.scrollable}

```{r annual_total, echo=TRUE}
# calculate annual total streamflow
msagel_annual <- msagel %>%
  group_by(wy, sim) %>%
  dplyr::summarize(str = sum(str), obs = sum(obs), minstr=min(str), minobs=min(obs))

# added a column to indicate if the parameter set is acceptable
msagel_annual = msagel_annual %>% mutate(accept = ifelse(sim %in% results_acc$sim, "Accept", "Not"))

# also add a column for class good parameters
msagel_annual$othercal = ifelse(msagel_annual$sim %in% other_cal_parms, "Accept B", "Not B")

```

## Plot

```{r annual_total_plot, echo=TRUE}
# plot annual total streamflow
  p4 = ggplot(msagel_annual, aes(x=factor(wy), y=minstr, 
  fill=interaction(accept, othercal))) +
               geom_boxplot(position = position_dodge(width = 0.75)) + labs(x = "Water Year", y = "Streamflow (mm)", fill = "Accept × OtherCal") +
  theme_minimal() + geom_point(aes(x=factor(wy), y=minobs), col="red")
  p4
```
